---
title: "13. Base R"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

```{r, include=FALSE, message =FALSE}
source("start.R")


food <- read_csv(file.path("data", "food.csv"))
food_prices <- read_csv(file.path("data", "food_prices.csv"))
```

## Introduction

The majority of the functions and objects that have been introduced in
throughout the first two parts of this text are provided by user-contributed
packages. Most of these come from a small set of packages that together
comprise a version of R code know as the "tidyverse". Included in this
set of packages are **dplyr**, **tidyr**, **readr**, and **ggplot2**.
Benefits of using these functions include more consistent naming conventions,
better documentations, and the fact that they are often built to express
theoretical models for data analysis (for example, relational database
techniques encoded in **dplyr** and the Grammar of Graphics created in the
**ggplot2** package). Downsides include their computational inefficency and
difficulties arising from using and debuging them within R packages.

There are a lot of strong opinions both for and against the tidyverse
universe of packages. We will avoid a long-winded discussion of these here;
links to some of the back-story are given in the reference section. As should
be clear at this point, this text has been written with the opinion that the
tidyverse suite of packages provide a great way to do data analysis and an
ideal starting point for learning data science in R. However, eventually it
will be useful to learn the underlying built-methods available within the
R programming language.

The functions available directly from R without loading any third-part packages,
commonly known as "base R", will become particularly important as we learn
how to do more complex programming and data scraping within this part of the
book. In this chapter we will re-start from the very basics by describing the
fundamental data types and objects within R as well. These topics will be made
easier by the fact that we have seen many of them indirectly in the preceeding
chapters. We will also provide an overview of introductory computer science
concepts such as control flow and function calling. The material is intended to
paced for readers who had no prior programming experience before working
through this text.

## Vectors

A vector is the most basic object type in R. It consists of an ordered sequence of
values that all have the same type. For example, a vector can contain a sequence of
numbers or a sequence of strings, but cannot contain a sequence that contains both
numbers and strings. To create a vector, we can use the function `c` (combine or
concatinate). Here is an example of a vector consisting of 6 numbers:

```{r}
vector_example <- c(1, 2, 3, 4, 10, 20)
```

You should recognize the `c` notation in our input to several tidyverse functions,
whenever we need to pass multiple values to a single argument (such as the limits
argument to `scale_x_continuous()`). You may object to the idea that a vector is
the most basic object type in R by pointing out that it seems that a single number
or string is actually the basic building-block In almost any other programming
language you would be correct, however R has a strange quirk in that individual
numbers and strings are actually vectors with length one.

Arithmetic operations can be applied to a vector. For example, adding the number
1 to our example vector will add one to each element:

```{r}
vector_example + 1
```

As with all of the tidyverse functions, this does not change the values `vector_example`.
If we want to save the changes we would need to explicitely create a new variable with
the `<-` operator.

We can also perform arithmetic operations between vectors of the same length by performing
the operation *component-wise*. That is, the elements of the two vectors are lined up with
one another and the operation(s) are performed on each pair. For example, adding our example
vector to itself will produce an output that doubles each component:

```{r}
vector_example + vector_example
```

Logical statements can also be applied to vectors; this creates a new vector of a special
**logical** type consisting of only `TRUE` and `FALSE` values. Here, we create a vector
showing which values are greater than 4:

```{r}
vector_example > 4
```

Finally, vectors can have an option set of names associated with each element. In order to
set these names, we use the function `names` and assign it a new vector of names that are
the same length as the input vector. Here we will assign the first 6 letters of the alphabet
as the names to the example vector:

```{r}
names(vector_example) <- c("a", "b", "c", "d", "e", "f")
```

Printing a named vector in R will show the names along with the values.

```{r}
vector_example
```

Logical vectors and names will be important for selecting and changing individual
parts of a vector, something that we will see in the following sections.

## Data type and lists

In the previous section we noted that vectors must consist of values that are
all the same type of data. We can check the type of data stored in a vector by
using the `class` function. For example, here we see that a vector of numbers
is of a "numeric" class:

```{r}
class(c(1, 2, 3, 4, 5))
```

And a vector of letters has a class called "character":

```{r}
class(c("A", "B", "C"))
```

Creating a logical condition creates a vector of a "logical" class:

```{r}
class(c(1, 2, 3, 4, 5) <= 3)
```

Data types are important as they can effect the kinds of operations
we are allowed to perform on a vector. For example, we will get an error
trying to add a number to a character vector.

```{r, error=TRUE}
class(c("A", "B", "C") + 1)
```

The concept of data types and the respective names should seem familiar as
we used the same idea when describing the types of data stored in the columns
of a tabular dataset.

R has a special type of vector called a "list". Each element of a list can store
an arbitrary object. In order to handle this flexibility, some of the semantics
for working with lists are different than those for other vectors. To create a
list, we can using the function `list` much like we used the function `c` above.

```{r}
list_example <- list(1, 2, "a", "b", 2 + 2 > 4)
list_example
```

Notice that it was possible to create a list that has elements that appear to
contain different types of data.

## Selecting and Modifying Vectors

In this section, we will see how to select a subset of a vector and then how
to use this selection to change some of the elements within an existing vector.
To start, let's reset with a named example vector:

```{r}
vector_example <- c(1, 2, 3, 4, 10, 20)
names(vector_example) <- c("a", "b", "c", "d", "e", "f")
```

There are four typical ways that we can take a subset of a vector. All of these
consist in placing square brackets at the end of the vector and putting a new
vector inside of the bracket. If the vector consists of positive integers, R
will extract elements of the vector *by position*. For example, passing a
vector of the numbers 1 and 5 will return the first and fifth elements:

```{r}
vector_example[c(1, 5)]
```

Recall that all numbers are vectors of length one, so we can use the same
method (i.e., `vector_example[1]`) to get a single value from a vector.

Likewise, passing a vector of negative integers will return a vector where
the corresponding positions have been removed:

```{r}
vector_example[c(-1, -5)]
```

If the vector has names, we can also pass a character vector of names. The
elements of the vector with the corresponding names are returned. Note that
if the vector has duplicated names, only the first element with a given
name will be returned.

```{r}
vector_example[c("a", "c", "f")]
```

Finally, we can also pass a logical vector of the same length as the target
vector. Elements corresponding with the value `TRUE` are returned. This is
usually used when the logical vector is created relative to the original vector.
For example, here is an quick way to extract the elements that are greater than
4:

```{r}
vector_example[vector_example > 4]
```

Changing a limited set of values within a vector uses the same notation as
selection, but requires putting the selection on left hand side of the
assignment arrow. For example, here we change the second element in our vector
to 100:

```{r}
vector_example[2] <- 100
vector_example
```

As mentioned above, lists are a special type of vector that itself contains
arbitrary objects. All of the methods described for selecting and assinging
vectors works for lists as well. However, note that these will always return
a list. For example, the code here returns a list of length one containing the
original list's second element, rather than the contained vector itself:

```{r}
list_example <- list(c(1, 2, 3), c(4, 5, 6), "a")
list_example[2]
```

There are two special ways of accessing the actual element within an individual
list component. First, we can use double square brackets with a single number
(note that these are also used in the output of the list to indicate each element).
So, to get the second element of the list as a numeric vector, we can use this
notation:

```{r}
list_example[[2]]
```

Alternatively, if a list has names, we can extract an individual value by using
the dollar sign operator (`$`) followed by the element's name:

```{r}
names(list_example) <- c("A", "B", "C")
list_example$B
```

Both of these techniques can also be used to assign new and existing values
to a list. The following code, for example, adds a new element to the list
named "new":

```{r}
list_example$new <- c("hello", "again")
list_example
```

Trying to extract an element from a list that does not exist will return a
special value called `NULL` rather than producing an error. This is a common
feature of lists that is employed in advanced R code.

## Matrix

In R, a matrix is a type of object consisting of a rectangular arrangement of
values of the same type. Most commonly matricies are composed of numbers, but
it is also possible to create other types of matricies (such as a matrix of
characters or logical values). One way to create a matrix is by using the
`matrix` function, which takes a vector of values and an indication of how
many rows (`nrow`) and/or columns (`ncol`) should be produced. Printing a
matrix in R will display the values as a rectangular grid:

```{r}
matrix_example <- matrix(c(1, 2, 3, 4, 5, 6), ncol = 3)
matrix_example
```

A matrix can be thought of as 2-dimensional extension of a vector.
As with vectors, we can perform mathematical operations component-wise:

```{r}
matrix_example + 1
```

Or by lining up values from matricies of the same shape:

```{r}
matrix_example + matrix_example
```

Selecting rows and columns of a matrix also works similarly to that
of a vector. We use square brackets and pass vectors to describe which
rows and columns to select. The row selection is done first, followed by
a comma, and the column selection. Rows and columns can be selected by
component identifiers, by negative component identifiers, by names, or
by passing a logical vector. Here is one way to select the first two
rows and columns from the matrix example:

```{r}
matrix_example[c(1, 2), c(1, 2)]
```

As a shorthand, we can select all rows by leaving the first
selection empty. The same method works for selecting all columns.

```{r}
matrix_example[, c(1, 2)]
```

The functions `colnames` and `rownames` can be used to set names of
the two dimensions in order to later use them for selection:

```{r}
colnames(matrix_example) <- c("a", "b", "c")
rownames(matrix_example) <- c("A", "B")

matrix_example
```

A quirk of R's notation for selecting subsets of a matrix occurs when
the resulting output has only one row or one column. In this case, R
will silently convert the single row or column back into a vector.

```{r}
matrix_example[, 2]
```

More often than not, this is the desired behavior, but occasionally this
can cause unexpected bugs. This is particularly true when building R
packages that need to handle a diverse set of inputs in a consistant way.
In order to prevent this behavior, add the option `drop = FALSE` (with
an additional comma) to the selection of the matrix.

```{r}
matrix_example[c(1, 2), 2, drop = FALSE]
```

The resulting output matrix now retains it two-dimensional shape.

## Control flow

Our application of tidyverse functions generally assumed that each call to
a component (i.e., `select`, `ggplot`, or `read_csv`) would be run exactly
once. We may re-run a chunk of code after modifying something, but at the
end of the day, each piece is run in a specific order and does not need to
be repeated. When writing code in Base R, as will most programming languages,
there is often the need to have more explicit control to repeat and
selectively apply lines of code. We will see several ways of doing this
in R.

As an example, consider the following set of code that creates a vector and
then adds the number 1 to the first component, the number 2 to the second
component, and so forth.

```{r}
vector_example <- c(10, 20, 30, 40)

vector_example[1] <- vector_example[1] + 1
vector_example[2] <- vector_example[2] + 2
vector_example[3] <- vector_example[3] + 3
vector_example[4] <- vector_example[4] + 4

vector_example
```

There are a number of ways of simplifying this code using the methods learned
above for manipulating vectors. Another approach that is more generalizable to
other more complex tasks is to write one line of code and indicate to R that
we want this line repeated for each component in our vector. This is possible
with a *for loop*, which has the following syntax in R:

```{r}
vector_example <- c(10, 20, 30, 40)

for (i in c(1, 2, 3, 4))
{
  vector_example[i] <- vector_example[i] + i
}

vector_example
```

The middle line of code is repeated four times with different values of the
variable `i`.

Within a for loop, we may want to perform a particular task only if a specific
condition is met. For example, we may want to add i to the i'th position only
if i is bigger than 2. To do this, we can use an *if statement*. The containing
code only gets executed if the condition is `TRUE`. Here is an example of
only adding i for the second two components of our vector:

```{r}
vector_example <- c(10, 20, 30, 40)

for (i in c(1, 2, 3, 4))
{
  if (i > 2)
  {
     vector_example[i] <- vector_example[i] + i
  }
}

vector_example
```

Finally, we may want to do something else the case that our condition is
FALSE. To do this, we can add on an *else statement*. Here, we reset the
first two elements to 1 and 2 respectively:

```{r}
vector_example <- c(10, 20, 30, 40)

for (i in c(1, 2, 3, 4))
{
  if (i >= 3)
  {
    vector_example[i] <- vector_example[i] + i
  } else {
    vector_example[i] <- i
  }
}

vector_example
```

Another way to repeat a set of code multiple times is to write a custom
function. To create a function in R we use the `function` function, and
assign this output to a new object name. We have to describe the names
of the function inputs, what happens to the inputs, and then provide a
call to the `return` function to describe what comes as the function's
output. Here is a small example that adds one to its input and returns
the result:

```{r}
add_one <- function(input)
{
  output <- input + 1
  return(output)
}

add_one(30)
```

As we have seen, function in R can have arguments with default values.
To define these, set them equal to a default value when creating the
function. Here, for example, is a function that adds two numbers together,
but sets the second number to 1 by default:

```{r}
add_something <- function(input, something = 1)
{
  output <- input + something
  return(output)
}

c(add_something(30), add_something(30, 4))
```

Almost all of the functions that we have seen in R are themselves written
using this exact same method.

## Apply

Apply functions are a final type of control flow is particularly popular in
R code and demands some particular attention. Let's take an example list
object with three elements:

```{r}
list_example <- list(c(1, 2, 3), seq_len(10), seq_len(4))
list_example
```

If we wanted to add one to each of the vectors inside of this list, it is
possible to use a for loop over each of the elements and the selection
notation introduced at the start of this chapter. As an alternative, we
can use the function `lapply` to apply a function to each element of a
list. Its first argument is the list and the second argument is a function
that we want to apply to each element. Here, we can apply the custom function
`add_one` to each element of our list:

```{r}
lapply(list_example, add_one)
```

Applying `lapply` to a non-list object will first turn the object into a list
and then apply the function to each element.

There is also a function called `apply` that applies a function to each row
or column in a matrix. Let's create a small matrix to test this out with:

```{r}
matrix_example <- matrix(seq_len(21), ncol = 3)
matrix_example
```

The apply function's first arugment is the matrix input. The second argument
is either the number `1`, corresponding to applying the rows of the matrix, or
the number `2`, for applying over the columns. The final argument is the function
that should be applied. Here is an example of applying a function over the rows
of a matrix:

```{r}
apply(matrix_example, 1, sum)
```

And here is another example applying functions over the columns of a matrix:

```{r}
apply(matrix_example, 2, sum)
```

Applying a function that returns vectors of a fixed length over a matrix will
return a new matrix:

```{r}
apply(matrix_example, 2, add_one)
```

While less useful, we can also apply a function that returns variable length
vectors. These will be returned as a list.
