---
title: "11. Application: Movies Dataset I"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

```{r, include=FALSE, message=FALSE}
source("start.R")
```

## Movies Data

Over the next two classes we will be working with a dataset of movies I have
constructed consisting of the top 100 grossing films for each year from 1970 to
2019. The data comes from IMDb. Today we will focus on getting familiar with
the various components of the data. Let's read in the four tables of data,
as well as a data dictionary, and then go through each of the tables.

```{r, message=FALSE}
movies <- read_csv(file.path("data", "movies_50_years.csv"))
m_genre <- read_csv(file.path("data", "movies_50_years_genre.csv"))
m_people <- read_csv(file.path("data", "movies_50_years_people.csv"))
m_dict <- read_csv(file.path("data", "movies_50_years_data_dictionary.csv"))
m_color <- read_csv(file.path("data", "movies_50_years_color.csv"))
```

The movies dataset contains one row for each movie. Most of the variables are
fairly straightforward; you can see the units by opening the data dictionary.
Three variables concern the movie's theatrical poster. These given the average
brightness (average pixel intensity), saturation (are the colors bright or
washed out / white), and complexity (a cartoon image would have a low
complexity; lots of text or icons would have a high complexity).

```{r, message=FALSE}
movies
```

A second dataset gives more detailed information about each poster by indicating
how much of a poster is of a certain color. If you want to look at the movie
poster itself, just search for a film on [IMDb](https://www.imdb.com) and
search for the film. The poster is the first image on the film's page.

```{r, message=FALSE}
m_color
```

We also have a dataset of movie genres. The data structure is straightforward,
but needs to be kept in its own table because a single movies can be assigned
to multiple genres.

```{r, message=FALSE}
m_genre
```

Finally, we also have a dataset of people associated with each film. We do not
have a lot of metadata about the people, but I have added a prediction of each
person's gender based on U.S. Social Security records. These are not always
correct (there is a confidence score included as well) but are useful for some
aggregate analyses.

```{r, message=FALSE}
m_people
```

There is a lot to unpack with these datasets. We will need to make use of the
many methods we have learned so far this semester to make sense of the data.
