---
title: "17. XML Web Scraping"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

```{r, include=FALSE, message=FALSE}
source("start.R")
source("cache.R")

library(xml2)
```

## CNN Lite: Iteration

Let finish the notes we covered in class to grab all of the pages from CNN.
To start, recreate the top stories data:

```{r}
url_str <- modify_url("https://lite.cnn.com/en")
res <- http_cache_get(url_str, cache_dir = "cache")
obj <- content(res, type = "text/html", encoding = "UTF-8")

top_stories <- tibble(
  title = xml_text(xml_find_all(obj, ".//li")),
  link = xml_attr(xml_find_all(obj, ".//li/a"), "href")
)

top_stories
```

To speed things up, we will use only the first 10 stories but the code can be
easily changed to include all of them. Note that here we need to collect two
different tables, one for metadata and one for stories. The unique id created
above will help if we later want to put these tables together.

```{r}
n <- 10

output_meta <- vector("list", n)
output_para <- vector("list", n)
for (i in seq_len(n))
{
  # grab the story
  url_str <- modify_url("https://lite.cnn.com/", path = top_stories$link[i])
  res <- http_cache_get(url_str, cache_dir = "cache")
  obj <- content(res, type = "text/html", encoding = "UTF-8")

  # create id
  id <- stri_extract(top_stories$link[i], regex = "/[^/]+\\Z")
  id <- stri_sub(id, 2, -1)
  
  # create and store metadata 
  output_meta[[i]] <- tibble(
    title = xml_text(xml_find_all(obj, ".//h2")),
    byline = xml_text(xml_find_first(obj, ".//p[@id='byline']")),
    published = xml_text(xml_find_first(obj, ".//p[@id='published datetime']")),
    source = xml_text(xml_find_first(obj, ".//p[@id='source']")),
    note = xml_text(xml_find_first(obj, ".//p[@id='editorsNote']"))
  )
  
  # create and story story text
  output_para[[i]] <- tibble(
    id = id,
    text = xml_text(xml_find_all(obj, ".//div[not(@style)]/p[not(@id)]"))
  )
}

cnn_meta <- bind_rows(output_meta)
cnn_text <- bind_rows(output_para)
```

Which yields a metadata table:

```{r}
cnn_meta
```

And a table with all of the story text:

```{r}
cnn_text
```





