---
title: "08. Creating Features"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)
library(stringi)

theme_set(theme_minimal())
options(pillar.min_character_chars = 15)
options(dplyr.summarise.inform = FALSE)
options(readr.show_col_types = FALSE)
options(ggrepel.max.overlaps = Inf)
options(width = 85L)

food <- read_csv(file.path("data", "food.csv"))
food_prices <- read_csv(file.path("data", "food_prices.csv"))
```

## Mutate verb

The final core **dplyr** verb that we will look at is used to create a new
feature in our data set based on other features that are already present.
This verb is called `mutate`, and works by giving it the name of the feature
you want to create followed by the code that describes how to construct the
feature in terms of the rest of the data.

As an example, consider computing the number of calories in an 200g portion of
each food. All of the features in the data set are currently given as 100g
portions, so to compute this we need to multiply the `calories` feature by 2.
To do this, we use the `mutate` verb to name and describe a new feature
`calories_200g`.

```{r}
food %>%
  mutate(calories_200g = calories * 2)
```

Notice that there is a new feature named `calories_200g` that has been added
as the last column in the data set. Because it is added at the end of the
data set, it gets hidden in the output shown above. Making use of `select`
allows us to see the new values:

```{r}
food %>%
  mutate(calories_200g = calories * 2) %>%
  select(item, food_group, calories, calories_200g)
```

And now we can see that the new column has been created by doubling the number
given the `calories` column.

Note that `mutate` can also be used to modify any existing column in the
data set by using the name of an extant feature. In this case the position of
the feature within the tables does not change.

The mutate verb itself has a relatively straightforward syntax. The main
challenge is knowing how to apply and chain together the various
transformations that are useful within an analysis. In the next few sections,
we highlight several common types of operations that we will be useful in
subsequent applications.

## Conditional values

Many of the uses for the mutate verb involve assigning one value when a set of
conditions is true and another if the conditions are false. For example,
consider creating a new feature called `sugar_level` based on the relative
amount of sugar in each food item. We might classify a food has having a "high"
sugar level if has more than 10g of sugar per 100g serving, and a "normal"
amount otherwise. In order to create this feature, we need the function
`if_else`.

The `if_else` function has three parts: a TRUE/FALSE statement, the value to
use when the statement is true, and the value to use when it is false. Here is
an example to create our new feature:

```{r}
food %>%
  mutate(sugar_level = if_else(sugar > 10, "high", "normal")) %>%
  select(item, food_group, sugar, sugar_level)
```

Looking at the first rows of data, we see that apples and bananas are classified
as high sugar foods, whereas the other sugar levels are given the sugar level
category of "normal".

The `if_else` function can be used to produce any number of categories by using
it multiple times. Let's modify our sugar level feature to now have three
categories: "high" (over 10g), "low" (less than 1g), and "normal" (between 1g
and 10g). There are several different ways to get to the same result, but I
find the easiest is to start by assigning a default value and then changing the
value of the new feature in sequence. For example, here some code that produces
our new categories:

```{r}
food %>%
  mutate(sugar_level = "default") %>%
  mutate(sugar_level = if_else(sugar < 1, "low", sugar_level)) %>%
  mutate(sugar_level = if_else(sugar > 10, "high", sugar_level)) %>%
  mutate(sugar_level = if_else(between(sugar, 1, 10), "normal", sugar_level)) %>%
  select(item, food_group, sugar, sugar_level)
```

In each `if_else` step we are telling the mutate function that if the condition
is false set `sugar_level` equal to itself. In other words, if the condition
does not hold, do not change the value of the feature.

In may wonder why we created a "default" value for the feature `sugar_level`.
It would have been one less line of code to set the default value to "normal"
and remove the final mutate function. The reason for the approach above is
three-fold. First, it's easier to understand what the code is doing in it's
current format because each condition ("high", "normal", and "low") is
explicitly coded. Secondly, it creates a nice check on our code and data. If we
find a row of the output that still has the value "default" we will know that
there is a problem somewhere. Finally, the code above will more safely handle
the issues with missing values, and issue that we will return to shortly.

## Creating labels

Another common type of manipulation that is used with the mutate verb is the
creation of a textual label. This can be done with the function `stri_paste`. It
takes any number of inputs and combines them into a single string. We can mix
both features in the data set as well as fixed strings to create useful labels
for data visualizations. For example, if we create a label feature describing
each fruit and its sugar content:

```{r}
food %>%
  mutate(label = stri_paste(item, " (", sugar, "g)")) %>%
  select(item, sugar, label)
```

The `label` feature could then be used as a textual label in a subsequent plot
with textual data. 

## Factors

R has a special data type called a "factor" (abbreviated "fct") that is
specifically designed to handle categorical features. It is typically not a
good idea to store data as a factor because the resulting features have some
odd, error-producing, behaviors. However, it can be useful to create a factor
as part of a mutate function just prior to creating a data visualizations.

For us, biggest difference between factors and character vectors is that a
factor vector has a default ordered of its unique values, called the factor's
"levels". Creating and understanding factors is useful because it allows us to
change the ordering of categories within visualizations and models (which by
default is done alphabetically).

One of the easiest ways to produce a factor feature with a given order is
through the function `fct_inorder`. It will order the categories in the same
order that they (first) appear in the data set. Combining this with the
`arrange` function provides a lot of control over how categories become ordered.
For example, the following code produces a bar plot of the food groups in our
data set arranged from the largest category to the smallest category:

```{r}
food %>%
  group_by(food_group) %>%
  summarize(n = n()) %>%
  arrange(desc(n)) %>%
  mutate(food_group = fct_inorder(food_group)) %>%
  ggplot() +
    geom_col(aes(food_group, n))
```

Other useful functions for manipulating categories include `fct_relevel` for
manually putting one category first and `fct_lump_n` for combining together
the smallest categories into a collective "Other" category.

## Mutate summaries

All of summary functions that were introduced in the previous notebook can also
be applied within the mutate version. Instead of reducing the data to a single
summary row, summarizing within the mutate verb duplicates the summary statistic
in each row of the data set. Here is an example of including the average number
of calories across all rows of the data set:

```{r}
food %>%
  mutate(calories_mean = mean(calories))
```

As with any call to mutate, all of the original features are kept in the output
and the new feature is added at the end. Using `select` we can verify that the
average calories has in fact been added to each row of the table.

```{r}
food %>%
  mutate(calories_mean = mean(calories)) %>%
  select(item, food_group, calories, calories_mean)
```

The power of mutate summaries becomes particularly clear when grouping the
data. If we group the data set by one or more features and apply a summary
function within a mutation, the repeated summaries will be done within
each group. Here is an example of adding the average calories of each food
group to the data set:

```{r}
food %>%
  group_by(food_group) %>%
  mutate(calories_mean = mean(calories)) %>%
  select(item, food_group, calories, calories_mean)
```

Following this with a filter, for example, would allow us to select all of the
foods that have a less than average number of calories within their food group.
We will see many examples of grouped mutate summaries throughout our
applications.

## Labels and themes

We have seen a number of ways to create and modify data visualizations. One
thing that we did not cover was how to label our axes. While many data
visualization guides like to stress the importance of labelling axes, while in
the exploratory phase of analysis it is often best to simply use the default
labels provided by R. These are useful for a number of reasons. First, they
require minimal effort and make it easy to tweak axes, features, and other
settings without spending time tweaking with the labels. Secondly, the default
labels use the feature names in our dataset. When writing code this is
*exactly* what we need to know about a feature to use it in additional plots,
models, and data manipulation tasks. Of course, once we want to present our
results to others, it is essential to provide more detailed descriptions
of the axes and legends in our plot. Fortunately, this is relatively easy using
the grammar of graphics.

In order to change the labels in a plot, we can use the `labs` function as an
extra part of our plot. Inside the function, we assign labels to the names of
aes values that you want to describe. Leaving a value unspecified
will keep the default value in place. Labels for the x-axis and y-axis will be
go on the sides of the plot. Labels for other aesthetics such as size and color
will be placed in the legend. Here is an example of a scatterplot with labels
for the three aesthetics:

```{r, message=FALSE}
food %>%
  ggplot() +
    geom_point(aes(x = calories, y = sat_fat, color = food_group)) +
    labs(
      x = "Calories per 100g",
      y = "Saturated Fat (grams per 100g)",
      color = "Food Group"
    )
```

Notice that the descriptions inside of the `labs` function is fairly long. The
code here breaks it up by putting each argument on its own line (indented a
further two spaces). This is good practice when using functions with a lot of
arguments.

We can also had a title (and optional subtitle and caption) to the plot by
adding these as named arguments to the `labs` function.

```{r}
food %>%
  ggplot() +
    geom_point(aes(x = calories, y = sat_fat, color = food_group)) +
    labs(
      title = "Main title",
      subtitle = "A little more detail",
      caption = "Perhaps the source of the data?"
    )
```

Another way to prepare our graphics for publication is to modify the *theme* of
a plot. Themes effect the way that the plot elements such as the axis labels,
ticks, and background look. By default through this book, we have set the
default plot to `theme_minimal`. I think that this is a great choice for
exploration of a dataset. As the name implies, it removes most of the clutter
of other choices while keeping grid lines and other visual cues to help
interpret a dataset. When presenting information for external publication, I
prefer to use the theme called `theme_sm` based on the work of Edward Tufte.
To set the theme, just call the following line of code sometime before making
your plot:

```{r}
theme_set(theme_sm())
```

Now, when we construct a plot it will use the newly assigned theme:

```{r}
food %>%
  ggplot() +
    geom_point(aes(x = sugar, y = total_fat))
```

The Tufte theme is designed to use as little "ink" as possible, thus focusing
the reader on the data. It can be a bit too minimal when first exploring the
dataset, but is a great tool for presenting your results.
