---
title: "02. Working with R and RMarkdown"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

```{r, include=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)

theme_set(theme_minimal())
options(pillar.min_character_chars = 15)
```

## R Notebooks

This file is called an R Notebook. It is a mixture of text (like this)
written in a format called *markdown*, and blocks of code that look
like this:

```{r}
2 + 2
```

You can run code a block of code by click on the green arrow in the top-left
of the code block. Try this on the block above; you should see the result of
the addition show up as a result below the code.

Most of the notes this semester will be given in the form of a new notebook.
During class we will work through the notes and embedded questions. Depending
on the pace of the day, we may not finishing all of the questions in the
notebook. You should try to finishing the remaining questions for homework
before the next class meeting.

Today we will cover some of the basics of running R code and introducing the
various parts of the RCloud software.

## Running R Code

We now want to give a very brief overview of how to run R code. We
will now only show snippets of R code and the output rather than a
screen shot of the entire RStudio session. Though, know that you
should think of each of the snippets as occurring inside of one of
the grey boxes in an RMarkdown file.

In one of its most basic forms, R can be used as a fancy calculator. We already
saw this above. Or, for example, we can divide 12 by 4:

```{r}
12 / 4
```

We can also store values by creating new *objects* within R. To do
this, use the `<-` (arrow) symbol. For example, we can create a new
object called `mynum` with a value of `8` by:

```{r}
mynum <- 3 + 5
```

Notice that the number will also show up in the upper left hand corner
of the RStudio window. We can now use our new object `mynum` exactly the
same way that we we would use the number 8. For example, adding it to 1
to get the number nine:

```{r}
mynum + 1
```

As we work through course notes this semester, you will find questions
scattered throughout the notebook. Try to answer these are we go along;
occasionally I may call on someone to show and/or tell their answers.
Here's our first question. In the code block below, divide the number 57
by 3:

```{r}
57 / 3
```

Some questions will also be in the form of a short text response, usually
followed with the word Answer, where you should put your response, like
below.

Is 57 a prime number? **Answer**: No, because it is evenly divisible by 3.

## Running functions

A function in R is something that takes a number of input values and
returns an output value. Generally, a function will look something like
this (Note: This is just an example. If you run the code you will see
an error because it is not a real function.):

```{r, eval=FALSE}
function_name(arg1 = input1, arg2 = input2)
```

Where `arg1` and `arg2` are the names of the inputs to the function (they
are fixed) and `input1` and `input2` are the values that we will assign
to them. The number of arguments is not always two, however. There may be any
number of arguments, including zero. Also, there may be additional optional
arguments that have default values that can be modified.

Let us look at an example function: `seq`. This function returns a sequence
of numbers. We will can give the function two input arguments: the starting
point `from` and the ending point `to`.

```{r}
seq(from = 1, to = 7)
```

The function returns a sequence of numbers starting from 1 and ending at
7 in increments of 1. Note that you can also pass arguments *by position*, in
which case we use the default ordering of the arguments. Here is the same code
but without the names:

```{r}
seq(1, 7)
```

There is also an optional argument `by` that controls the spacing between
each of the numbers. By default it is equal to 1, but we can change
it to spread the point out by half spaces.

```{r}
seq(from = 1, to = 7, by = 0.5)
```

We will learn how to use numerous functions throughout the semester.

## Loading Data

The goal of this course is to learn how to work with data. Not surprisingly,
we will learn a number of functions for loading data into R. In the next
class we will introduce more of the details around organizing and loading
data. Let's just see a quick example for now. We will load a data set called
`foods.csv` that is stored in the folder `data`. To start, we need to load
the appropriate R library. (Note: Usually we will do this at the start of the
notebook.)

```{r, message=FALSE}
library(tidyverse)
```

Next, the following code uses the function `read_csv` to load this data set into R,
save it as an object called `food`, and prints out the first several rows.

```{r, message=FALSE}
food <- read_csv("data/food.csv")
food
```

You should also notice that the food data set appears in the upper left-hand
corner of the screen. Clicking on it will open an Excel-like version of the
data set. We will be working with these food data as a source of examples
throughout the first few weeks of the class.

## Formatting

It is very important to properly format your code in a consistent way.
Even though the code may run without errors and produce the desired results,
you will make your life easier by writing well-formatted code from the start.
This makes it easier to read and debug in the future. We will follow the following
guidelines:

- always put one space before and after an equals sign or assignment arrow
- always put one space around mathematical operations (such as `+` and `*`)
- always put one space after a comma, but no space before a comma

In the code block below, I wrote some code that selects all of the
fruits and vegetables and produces a scatter plot with all the vegetables
in the data set. We will learn more about these functions code over the
next few weeks. For now, just reformat the code.

```{r}
veggie <- filter(food, food_group == "vegetable")
ggplot(veggie, aes(x = calories, y = total_fat)) + geom_point()
```

Try running the code to see how easily RStudio makes it to embed
visualizations into RMarkdown notebooks.

## Practice: Running R Code

That's all we have today for new material. In the rest of the
notebook are some further practice questions to see how well you
understood the material.

In the code block below, make a variable named `fav_number` and set it
equal to your favorite number. (Note, the quotes are not part of the
variable name).

```{r}
fav_number <- 8
```

In the code below, apply the function `log10` to your favorite
number. (Again, the quotes are not part of the function name).

```{r}
log10(fav_number)
```

Note that the only point of the exercises here are to familiarize
you with running code in R, creating variables, and applying functions.
We will get to more interesting tasks soon!
