---
title: "Notebook 07 -- Solutions"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---


## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)
library(stringi)

theme_set(theme_minimal())
options(pillar.min_character_chars = 15)
options(dplyr.summarise.inform = FALSE)
options(readr.show_col_types = FALSE)
options(ggrepel.max.overlaps = Inf)
options(width = 85L)
```

I have set the options `include=FALSE` and `message=FALSE` to avoid cluttering
the solutions with all the output from this code.

## Practice

### Movies data

For this notebook, we will be creating two datasets, reading them into R, and
producing several data visualizations.

For the first task, we will create a dataset as a class describing a set of
movies. Each student will be tasked with creating three rows of data. We
will record the following items in the GoogleSheets file linked to on the
class website:

- student number
- title
- date of release (year, month, and day)
- budget
- box-office numbers
- genre
- personal rating of the film, from 1-5 stars

Once you are finished, save the data set as a csv file, upload it to the "data"
directory in RStudio, and read it into R in the following block of code:

```{r, message=FALSE}
movies <- read_csv(file.path("data", "movie.csv"))
movies
```

Then, produce a scatterplot below with the budget on the x-axis, box-office
numbers on the y-axis, points colored by the year of release, and the size of
the points corresponding to your personal rating. Use a text repel layer to
label the movies by their title. Include axis labels and titles, along with
units of measurement.

```{r, warning=FALSE}
movies %>%
  ggplot(aes(budget, box_office)) +
    geom_point(aes(color = release_year, size = rating)) +
    geom_text_repel(aes(label = title), size = 3) +
    labs(x = "Budget (USD, Millions)", x = "Box Office (USD, Millions)") +
    scale_x_log10() +
    scale_y_log10() +
    scale_color_viridis_b()
```

If you think it is needed, also include logarithmic scales for the x- and
y-axes.

### Friends data

For the second data set, you are going to collect information about ten of your
friends on your own. You can do this in any software you would like. At a
minimum, record the following features:

- their name
- where you met them (make your own categories, such as "university",
"high school", "camp")
- the year you met the friend
- personal rating of the friend, from 1-5 stars

Once you are finished, save the data set as a csv file and read it into R in
the following block of code:

```{r, message=FALSE}
friends <- read_csv(file.path("data", "friends.csv"))
friends
```

Then, produce a scatter plot below using any two variables that you find
interesting as the x- and y-axes. Include a highlight layer showing a subset
of your friends. These could be friends you met at a certain time in your life,
highlighting your best friend(s) or something else that is interesting to you.
Include axis labels and titles, along with units of measurement.

```{r}
best_friends <- friends %>%
  filter(rating > 5)

friends %>%
  ggplot(aes(year_met, where_met)) +
    geom_point(size = 2) +
    geom_point(data = best_friends, color = "olivedrab", size = 2) +
    geom_text_repel(aes(label = name), data = best_friends, color = "olivedrab") +
    labs(x = "Year Met", y = "Where Met")
```
