---
title: "Notebook 19"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggrepel)
library(stringi)
library(sf)

source("cache.R")

theme_set(theme_minimal())
options(pillar.min_character_chars = 15)
options(readr.show_col_types = FALSE)
options(width = 77L)
options(dplyr.summarise.inform = FALSE)
options(ggrepel.max.overlaps = Inf)
Sys.setlocale(locale = "en_US.UTF-8")
sm_centroid <- function(data) {
  suppressWarnings({ z <- st_coordinates(st_centroid(data)) })
  return(tibble(lon = z[,1], lat = z[,2]))
}

theme_set(theme_minimal())
```

### Introduction

This notebook introduces methods for working with spatial data in R. We will see
methods for creating spatial data from a data set with longitude and latitude,
how to plot spatial data, how to project spatial data, and how to compute
spatial summaries from lines and polygons.

Note that I have set the figure height in some chunks to make the images easier
to read. Feel free to adjust based on your screen and operating system.

### Spatial Points

Let's start by reading in the familiar US cities population data. I will take 
just a single year of data and remove rows without location information.

```{r}
us <- read_csv("data/us_city_population.csv") %>%
  filter(year == 2010) %>% 
  filter(!is.na(lon), !is.na(lat))
```

Below, produce a scatter plot with each city plotted according to its longitude
and latitude.

```{r, question-01}
us %>% 
  ggplot(aes(lon, lat)) +
    geom_point()
```

The map above is not bad, we can do better by turning this into a "proper" 
geospatial object. The code below creates a spatial points dataset by indicating
that the columns "lon" and "lat" contain longitude and latitude records. The 
crs code (4326) indicates that these are "raw" longitude and latitude and the
remove flag indicates that we do not want to lose all the other columns in the
data. You can typically use these same values of these parameters in other
tasks.

```{r}
us_geo <- us %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE)
```

Print the `us_geo` data below and also look at in the data viewer. Take a few
moments to notice how it differs from other tabular datasets we have used this
semester.

```{r, question-02}
us_geo
```

Now, create a plot with `us_geo` using the geometry `geom_sf`. Do not set any
aesthetics; it will now what to do on its own! Take note of how this does and
does not differ from the previous plot above.

```{r, question-03}
us_geo %>% 
  ggplot() +
    geom_sf()
```

The x and y aesthetics do not need to be set for `geom_sf`, but we can set other
aesthetics just as with any set of points. For example, in the code below change
the points to be a different (fixed) color and size. Pick whatever size and 
color you would like.

```{r, question-04}
us_geo %>% 
  ggplot() +
    geom_sf(color = "pink", size = 5)
```

One nice thing that we can do with spatial data is change the projection before
plotting. This create a better representation of the curved earth on a flat plot
based on the region of the world that we are looking at. To change the projection,
just pipe the data to the function `st_transform()`. The function takes one 
argument, a numeric code giving the crs number of the projection. Above, you 
have already seen that 4326 is the projection for plain longitude and latitude.
Redo your plot of the data (with the default size and color) below, but use a
projection with a crs of 5069 (this goes before you pipe the data to the plot).
You should notice a very different look to the output.

```{r, question-05}
us_geo %>% 
  st_transform(5069) %>%
  ggplot() +
    geom_sf()
```

There are two other spatial geometries: `geom_sf_text` and `geom_sf_label`. They
add textual annotations to the plot based on the location of each row. As with
`geom_sf`, you do not need to set the x- and y-aesthetics. However, you do need
to specify the label aesthetic. Below, create a plot using the coordinate system
with crd 5096 which uses `geom_sf_text` (rather than `geom_sf`) to label the
cities. Use the city name as the label.

```{r, fig.height=8, question-06}
us_geo %>% 
  st_transform(5069) %>%
  ggplot() +
    geom_sf_text(aes(label = city))
```

### Spatial Lines

Spatial data can also associate lines with each row rather than a single point.
It is less likely that you will create this type of spatial yourself, instead
reading the data from a file that is already designed to store spatial
information. The file type we will use is called GeoJSON. It is a JSON file as
we saw in the previous unit, but it has a standard format so we do not need to
parse it ourselves. The function `read_sf` can read geojson, storing the data
in R as a spatial data frame. 

Let's read in a dataset of roads in New York City. Take another moment to look
at the data print out and the data in the data viewer.

```{r}
nyc <- read_sf("data/ny_roads.geojson")
nyc
```

Spatial lines can be plotted with `geom_sf` much like spatial points. As with
points, we can do this without specifying any x- or y-aesthetic. Try this here
with the `nyc` data; you should see a familiar shape of the city.

```{r, fig.height=8, question-07}
nyc %>%
  ggplot() +
    geom_sf()
```

And again, as with points, we can change the color and size of the lines if 
needed. Below, repeat your plot of the NYC roads in a color of your choosing.

```{r, fig.height=8, question-08}
nyc %>%
  ggplot() +
    geom_sf(color = "pink")
```

There are a number of functions that compute summaries of spatial data. For 
example if we type the following inside of a mutate verb:

  len = as.numeric(st_length(geometry))
  
We will get the length of a line geometry in metres. Try this below, sorting the
output from the longest to the shortest road in NYC:

```{r, fig.height=8, question-09}
nyc %>%
  mutate(len = as.numeric(st_length(geometry))) %>%
  arrange(desc(len))
```

Notice that some of the city roads have duplicates because there are multiple
segments of a road broken up by some other feature. 

Now, let's do something a bit more involved. In the block below, create a new
data set named `nyc_broad` that only contains segements of the street "Broadway". 

```{r, question-10}
nyc_broad <- nyc %>%
  filter(name == "Broadway")
```

Now, create a map of the streets in NYC in which most streets are the color 
"grey85" but Broadway is in the color "orange". Also make Broadway have a size
of 2. Note: this can be done by layering two `geom_sf` layers.

```{r, fig.height=8, question-11}
nyc %>%
  ggplot() +
    geom_sf(color = "grey85") +
    geom_sf(data = nyc_broad, color = "orange", size = 2)
```

### Spatial Polygons

Finally, let's look data consisting of spatial polygons in which each row is
associated with one or more polygonal regions. Spatial polygons can be stored
in GeoJSON and read with `read_sf`, just as with the spatial lines data. Let's
read in a data set of US state boundaries. Once again, take a moment to look at
the data.

```{r}
state <- read_sf("data/state.geojson")
state
```

As with spatial points and lines, we can create a reasonable plot by piping the
`state` data in to `ggplot` and using `geom_sf`. Try that here:

```{r, question-12}
state %>%
  ggplot() +
    geom_sf()
```

It is much more clear with polygons how distorted the default projection makes
everything. Modify the plot you did above in the block below using the
projection with CRS 5069.

```{r, question-13}
state %>%
  st_transform(5069) %>%
  ggplot() +
    geom_sf()
```

We can modify spatial data just as with any other dataset. Below, filter the 
states data to exclude Alaska ("AK") and Hawaii ("HI"). Continue to use the
5069 projection.

```{r, question-14}
state %>%
  filter(!(abb %in% c("AK", "HI"))) %>%
  st_transform(5069) %>%
  ggplot() +
    geom_sf()
```

With lines we were able to compute the length of the line using an summary
function. A similar approach can be used to determine the area of a polygon.
This code inside of a mutate function computes the area of the polygon in
square kilometers:

   area = as.numeric(st_area(geometry)) / 1e6
   
Below, compute this value for all of the states and order the output from the
smallest state to the largest state:

```{r, question-15}
state %>%
  mutate(area = as.numeric(st_area(geometry)) / 1e6) %>%
  arrange(area)
```

We can add some information to a spatial plot of polygons by using the fill 
aesthetic. Below, plot the states (removing Alaska and Hawaii) in projection
5069 with the fill aesthetic proportional to the area of the state.

```{r, question-16}
state %>%
  filter(!(abb %in% c("AK", "HI"))) %>%
  mutate(area = as.numeric(st_area(geometry)) / 1e6) %>%
  st_transform(5069) %>%
  ggplot() +
    geom_sf(aes(fill = area))
```

The default color scale is awful in general, and particularly bad for maps. 
Adjust the previous plot in the block below by adding the following fill scale:

  scale_fill_distiller(
    trans = "log2", palette = "Spectral", guide = "legend", n.breaks = 10
  )

```{r, question-17}
state %>%
  filter(!(abb %in% c("AK", "HI"))) %>%
  mutate(area = as.numeric(st_area(geometry)) / 1e6) %>%
  st_transform(5069) %>%
  ggplot() +
    geom_sf(aes(fill = area)) +
    scale_fill_distiller(
      trans = "log2", palette = "Spectral", guide = "legend", n.breaks = 10
    )
```

Finally, I wrote a custom function above called `sm_centroid` that automatically
add the centroid of spatial points to a data set. You can use it as follows:

```{r}
state %>%
  mutate(sm_centroid(geometry))
```

We will make use of this in several upcoming notebooks.

