---
title: "Notebook 03 -- Solutions"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)

theme_set(theme_minimal())

food <- read_csv(file.path("data", "food.csv"))
```

I have set the options `include=FALSE` and `message=FALSE` to avoid cluttering
the solutions with all the output from this code.

## Practice

### Cities Data

To further practice the Grammar of Graphics, we will work with the largest
cities data set.

```{r, message=FALSE}
cities <- read_csv(file.path("data", "largest_cities.csv"))
cities
```

Population numbers are given in millions and areas are given in square
kilometers. The three different kinds of populations refer to different
ways of describing what is included in a city. The variable `population` is a
measurement from the UN (2018) that tries to consistently define the region of
a city. More information is available in the U.N.'s World Urbanization
Prospectus, which is available
[here](https://population.un.org/wup/Publications/Files/WUP2018-Highlights.pdf).
The other three measurements refer to:

- **city**: city proper, defined by local administrative boundaries
- **metro**: according to UNICEF, the Metro Region is an area "comprising the
urban area as a whole and its primary commuter areas... In addition to the city
proper, a metropolitan area includes both the surrounding territory with urban
levels of residential density and some additional lower-density areas that are
adjacent to and linked to the city." See
[UNICEF](https://www.unicef.org/sowc2012/statistics.php) for more information.
- **urban**: an area that consists of a "continuously built up land mass of
urban development that is within a labor market (metropolitan area or
metropolitan region) and contains no rural land".

Several of the questions in this and the following notebooks concern the
analysis of the relationship between these various definitions.

### Scatterplot

In the code below, write the code to produce a scatter plot with population on
the x-axis and the city population on the y-axis. I have set `warning=FALSE`
because there are some missing data values and it will be helpful to turn off
the warnings about these data rows.

```{r, warning=FALSE}
cities %>%
  ggplot() +
    geom_point(aes(x = population, y = city_pop))
```

Now, in the code region below, produce a scatter plot with population on the
x-axis and metro population on the y-axis.

```{r, warning=FALSE}
cities %>%
  ggplot() +
    geom_point(aes(x = population, y = metro_pop))
```

Which of the two regional populations (metro or city) seem to more closely
relate to the U.N.'s definition of a cities population (e.g., the `population`
variable). **Answer**: The metropolitan population is much closer to the UN
definition than the city population variable.

## Text Plot

Produce a plot with population on the x-axis and metro population on the y-axis
and label each point with the city name.

```{r, warning=FALSE}
cities %>%
  ggplot() +
    geom_text(aes(x = population, y = metro_pop, label = city))
```

What two cities seem to have the largest (absolute) discrepancy between their
UN population and the metro population? **Answer**: Jakarta and Seoul seem to
stand out the most (their metro populations are much larger than the offical
UN population numbers).

Produce a plot with a text geometry where longitude (`lon`) is on the x-axis,
latitude is on the y-axis (`lat`), and the labels are given by the city names
(`name`).

```{r}
cities %>%
  ggplot() +
    geom_text(aes(x = lon, y = lat, label = name))
```

This should roughly resemble a map of the world. In the code below, produce
a modified version of this plot, replacing the text layer with two layers:
a text repel layer and well as a point layer.

```{r}
cities %>%
  ggplot() +
    geom_text_repel(aes(x = lon, y = lat, label = name)) +
    geom_point(aes(x = lon, y = lat))
```

Which of these two plots do you find easier to read, and why? (Note, there is
no clear correct answer here). **Answer**: I slightly prefer the second, but it
would be better to, perhaps, show only a subset of the labels. We will see how
to do this soon!
