---
title: "Notebook 19 -- Solutions"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)
library(stringi)
library(sf)
library(units)

theme_set(theme_minimal())
options(dplyr.summarise.inform = FALSE)
options(width = 77L)

sm_centroid <- function(data) {
  suppressWarnings({ z <- st_coordinates(st_centroid(data)) })
  return(tibble(lon = z[,1], lat = z[,2]))
}
```

## French COVID-19 Data

Here are the three French datasets that we will be working with. They contain
spatial information, population metadata, and coronavirus numbers at the level
of French *départements*. These are geographic areas that are an important
political entities in France. There are 101 départements; 96 in mainland Europe
and 5 overseas (called the *DOM*, or  Départements d'outre-mer).

```{r, message = FALSE}
dept <- read_sf(file.path("data", "france_departement.geojson"))
pop <- read_csv(file.path("data", "france_departement_population.csv"))
covid <- read_csv(file.path("data", "france_departement_covid.csv"))
```

The coronavirus data is stored with one row for each combination of day and
département. We have the cumulative number of people who died in each day from
COVID-19, the total number currently in hospital, the total number currently
in *reanimation* (this is similar to ICU, but not exactly equivalent, so I
used the french term here), and the cumulative number of newly recovered. Notice
that deceased and recovered are the *total* counts of people who have died or
recovered, whereas hospitalised and reanimation are the numbers at that moment
of patients in each group. There are columns indicating the number of new
hospitalisations and reanimations, but these have many missing data points.

## Practice

### Map projections

To get a better sense of how map projections work, slice the `dept` data to
include only rows 96-101. This gives one département in Europe and all of the
DOM. Plot the data using both `geom_sf` and `geom_sf_text` (for a label, use
`departement_name`). Do not project the data, so that it uses just longitude
and latitude. Note that Val-d'Oise is in Europe.

```{r, question-01, warning = FALSE}
dept %>%
  slice(96:101) %>%
  ggplot() +
    geom_sf() +
    geom_sf_text(aes(label = departement_name))
```

Take the code from above and use the EPSG-3943 projection; this was
recommend for use with European France. Notice that the European
part of the map has the lat- and lon- lines the closest to parallel to
the x- and y-axes.

```{r, question-02}
dept %>%
  st_transform(3943) %>%
  slice(96:101) %>%
  ggplot() +
    geom_sf() +
    geom_sf_text(aes(label = departement_name))
```

Finally, use the EPSG-5879 projection with the same data:

```{r, question-03}
dept %>%
  st_transform(5879) %>%
  slice(96:101) %>%
  ggplot() +
    geom_sf() +
    geom_sf_text(aes(label = departement_name))
```

Note how distorted the rest of the plot becomes, but how nice the lines look
near Mayotte and La Réunion.

### Population Density

Combine the methods in the notes to add a population density variable to the
dataset (people per square kilometer). Plot the data spatially using
color to show the population density for the first 96 rows of the dataset.
Use a viridis color scale and appropropriate projection.

```{r, question-04}
dept %>%
  mutate(area = as.numeric(set_units(st_area(geometry), "km^2"))) %>%
  left_join(pop, by = "departement") %>%
  mutate(density = population / area) %>%
  slice(1:96) %>%
  st_transform(3943) %>%
  ggplot() +
    geom_sf(aes(fill = density), color = "black", size = 0.1) +
    scale_fill_viridis_c()
```

You should see that Paris and the the surrounding areas are by far the most
dense areas. This is because they are the only départements that include only a
dense city area and not any of the surrounding countryside. To fix this we need
to be more careful about how we define a color palette.

Usually, we use color only as a secondary element in a plot. However, in a map
we often need to use color to show the main feature of interest. This means that
we have to be fairly careful about how colors are defined. Change the plot
from your previous question to have the following scale:

- `scale_fill_distiller(trans = "log2", palette = "Spectral", guide = "legend", n.breaks = 10)`

```{r, question-05}
dept %>%
  mutate(area = as.numeric(set_units(st_area(geometry), "km^2"))) %>%
  left_join(pop, by = "departement") %>%
  mutate(density = population / area) %>%
  slice(1:96) %>%
  st_transform(3943) %>%
  ggplot() +
    geom_sf(aes(fill = density), color = "black", size = 0.1) +
    scale_fill_distiller(trans = "log2", palette = "Spectral", guide = "legend", n.breaks = 10)
```

Other options for palettes include:

  - **Diverging:** BrBG, PiYG, PRGn, PuOr, RdBu, RdGy, RdYlBu, RdYlGn, Spectral
  - **Qualitative:** Accent, Dark2, Paired, Pastel1, Pastel2, Set1, Set2, Set3
  - **Sequential:** Blues, BuGn, BuPu, GnBu, Greens, Greys, Oranges, OrRd, PuBu,
                    PuBuGn, PuRd, Purples, RdPu, Reds, YlGn, YlGnBu, YlOrBr, YlOrRd

Keep this in mind as we continue to work with spatial data, as the defaults will
often look terrible and be hard to interpret.

### Creating spatial data

It is time to get a little bit more complicated. Let's read in a dataset of
the largest French cities:

```{r, message=FALSE}
cities <- read_csv(file.path("data", "france_cities.csv"))
cities
```

This is just a normal data frame object, but we can convert it into a
spatial object with the following code:

```{r}
cities <- st_as_sf(cities, coords = c("lon", "lat"), remove = FALSE)
st_crs(cities) <- 4326  # set the default projection; EPSG-4326 => lat/long
cities
```

Unlike the geometries in `dept`, each row here is only a single point rather
than a polygon. We can do spatial scatterplots without dealing with geometry
objects, but need to do this convertion to make projections work correctly.

In the code below, take the last plot you made and add a `geom_sf_label` layer
showing the names and locations of the largest 10 French cities. This helps add
context to the plot for anyone not particularly familiar with the locations of
the major cities (Paris is already easy to see; the others less-so).

```{r, question-06}
dept %>%
  mutate(area = as.numeric(set_units(st_area(geometry), "km^2"))) %>%
  left_join(pop, by = "departement") %>%
  mutate(density = population / area) %>%
  slice(1:96) %>%
  st_transform(3943) %>%
  ggplot() +
    geom_sf(aes(fill = density), color = "black", size = 0.1, alpha = 0.5) +
    scale_fill_distiller(trans = "log2", palette = "Spectral", guide = "legend", n.breaks = 10) +
    geom_sf_label(aes(label = city), data = slice(cities, 1:10))
```

You can adjust the number of cities based on how large the plot is on your screen.

### Integrating the COVID-19 data

Finally, let's integrate a bit of the COVID-19 data into our spatial analysis.
Start by creating a dataset called `covid_april` by filtering the data to
include only the data from "2020-04-30" and compute the number of COVID-19
deaths per 100k residents. This will require using the `pop` and `covid`
datasets.

```{r, question-07}
covid_april <- covid %>%
  filter(date == "2020-04-30") %>%
  left_join(pop, by = "departement") %>%
  mutate(morality_rate = deceased / population * 100000)
covid_april
```

For the 96 départements in Metropolitain France, draw a map showing the
number of people who have died from COVID-19 per 1000 residents on 30 April.
Consider using the code you had in the previous plot (with a distiller
palette and the top cities shown).

```{r, question-08}
# I added axis labels and titles; these were not required in the question
dept %>%
  mutate(area = as.numeric(set_units(st_area(geometry), "km^2"))) %>%
  left_join(covid_april, by = "departement") %>%
  slice(1:96) %>%
  st_transform(3943) %>%
  ggplot() +
    geom_sf(aes(fill = morality_rate), color = "black", size = 0.1, alpha = 0.5) +
    scale_fill_distiller(trans = "log2", palette = "Spectral", guide = "legend", n.breaks = 10) +
    geom_sf_label(aes(label = city), data = slice(cities, 1:10), size = 2) +
    labs(
      x = "", y = "", fill = "Deaths Per 100k People",
      title = "COVID-19 Mortality in France Métropolitaine",
      subtitle = "As of 30 April 2020",
      caption = "Data Source: Santé publique France and data.gouv.fr"
    ) +
    theme_void()
```

As a last step (you will find you can copy much of the previous code),
show the number of people who were hospitalised per 100k residents
on "2020-10-01".

```{r, question-09}
covid_now <- covid %>%
  filter(date == "2020-10-01") %>%
  left_join(pop, by = "departement") %>%
  mutate(hosp_rate = hospitalised / population * 100000)

dept %>%
  mutate(area = as.numeric(set_units(st_area(geometry), "km^2"))) %>%
  left_join(covid_now, by = "departement") %>%
  slice(1:96) %>%
  st_transform(3943) %>%
  ggplot() +
    geom_sf(aes(fill = hosp_rate), color = "black", size = 0.1, alpha = 0.5) +
    scale_fill_distiller(trans = "log2", palette = "Spectral", guide = "legend", n.breaks = 10) +
    geom_sf_label(aes(label = city), data = slice(cities, 1:10), size = 2) +
    labs(
      x = "", y = "", fill = "Rate",
      title = "COVID-19 Hospitalisation Rate (per 100k residents)",
      subtitle = "On 01 October 2020",
      caption = "Data Source: Santé publique France and data.gouv.fr"
    ) +
    theme_void()
```

If you have time remaining, consider adding some labels to the last two plots to
make them closer to something that you might include a published report. You
might consider adding `theme_void` to minimize the ink used on the axes.
