---
title: "Notebook 09 -- Solutions"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)
library(stringi)
library(broom)

theme_set(theme_minimal())
options(pillar.min_character_chars = 15)
options(dplyr.summarise.inform = FALSE)
options(readr.show_col_types = FALSE)
options(ggrepel.max.overlaps = Inf)
options(width = 85L)

food <- read_csv(file.path("data", "food.csv"))
food_prices <- read_csv(file.path("data", "food_prices.csv"))
```

I have set the options `include=FALSE` and `message=FALSE` to avoid cluttering
the solutions with all the output from this code.

# Practice

## Largest City and Köppen Datasets

In this notebook we will work with the largest cities dataset:

```{r, message=FALSE}
cities <- read_csv(file.path("data", "largest_cities.csv"))
```

We will also load a dataset containing metadata about the Köppen climate
classification scores.

```{r, message=FALSE}
koppen <- read_csv(file.path("data", "koppen_metadata.csv"))
koppen
```

## Joining Climate Data

In the code block below, use the `left_join` function to combine the `cities`
data set with the `koppen` metadata. Note that the key has a different name in
each of the two data sets.

```{r, question-01}
cities %>%
  left_join(koppen, by = c("koppen_code" = "code"))
```

Take a close look at the data set you created in the last question. Scrolling
all the way to the final columns, you should see the name of the climate region
and the default color added into the data set.

Something, however, has happened to the first variable in the combined
data set. What is the (new) name of the first variable in the combined
output? **Answer**: name.x

It turns out that both the `cities` data set and the `koppen` dataset both
contain a variable named "name". In R, a data set cannot contain two columns
with the same names. By default, `left_join` adds a suffix to distinguish
variables that came from the first data set and those that came from the
second dataset. I generally find the default suffix values to not be
ideal because they change both the names in the left table (the one that
contains the observations of interest) as well as the the new metadata
columns. Let's see how to change these suffix values.

The `left_join` function has an optional argument called `suffix`. You
can supply two string values to it to describe what suffix values are
placed on the first and second tables in the join. The default value is
`c(".x", ".y")`. I typically find something with the first entry empty
to be more practical. So, for example: `c("", "_koppen")`. In the code
below, repeat the left join with this suffix option:

```{r, question-02}
cities %>%
  left_join(koppen, by = c("koppen_code" = "code"), suffix = c("", "_koppen"))
```

Verify that the variable names match what you expected to find in the new
data set. The same suffix option can be used to modify the output for other
table joins.

Below, pipe the joined data set from the last question into a scatterplot with
longitude on the x-axis and latitude on the y-axis. Color the points based on
the default Köppen codes using an identity color scale:

```{r, question-03}
library(ggmaptile)

cities %>%
  left_join(koppen, by = c("koppen_code" = "code"), suffix = c("", "_koppen")) %>%
  ggplot(aes(lon, lat)) +
    stat_maptiles(zoom = 3, quiet = TRUE) +
    geom_point(aes(color = color), size = 3) +
    scale_color_identity() +
    theme_void()
```

### Manual colors with a legend

There is one problem with the plot that we created to color the
cities by their Köppen codes. There's no legend for the plot. A
way to rectify this problem is to color the points using the
variable `name_koppen` and use the colors in the `koppen` metadata
table to describe a manual scale with `scale_color_manual`.
Specifically, we need:

   `scale_color_manual(values = koppen$color, breaks = koppen$name)`.

Below, redo your plot with this color value and scale.

```{r, question-04}
cities %>%
  left_join(koppen, by = c("koppen_code" = "code"), suffix = c("", "_koppen")) %>%
  ggplot(aes(lon, lat)) +
    geom_point(aes(color = name_koppen)) +
    scale_color_manual(values = koppen$color, breaks = koppen$name)
```

The plot should like nice, but does requires a wide screen to see all of the
colors given the long descriptions in the legend.

## Creating join table

Sometimes it can be useful to create a small metadata table directly in R using
the `tibble` function. This can than be used to more easily do mutating or 
filtering of a dataset. We will practice this by looking at the continents in
the `cities` dataset. To start, compute the unique values of the continent 
variable in the dataset (there are many different ways to do this):

```{r, question-05}
cities %>%
  group_by(continent) %>%
  summarize()
```

Here are the continent names in French:

- Asie, Amérique du Sud, Amérique du Nord, Afrique, Europe)

In the next question, use the mutate function to create a new variable called
`continent_fr` in the `cities` data that gives the continent names in French.
Pipe this to a plot with longitude on the x-axis, latitude on the y-axis, and
the points colored to the name of the continent in French:

```{r, question-06}
cities %>%
  mutate(continent_fr = "default") %>%
  mutate(continent_fr = if_else(continent == "Asia", "Asie", continent_fr)) %>%
  mutate(continent_fr = if_else(continent == "South America", "Amérique du Sud", continent_fr)) %>%
  mutate(continent_fr = if_else(continent == "North America", "Amérique du Nord", continent_fr)) %>%
  mutate(continent_fr = if_else(continent == "Africa", "Afrique", continent_fr)) %>%
  mutate(continent_fr = if_else(continent == "Europe", "Europe", continent_fr)) %>%
  ggplot(aes(lon, lat)) +
    geom_point(aes(color = continent_fr))
```

Next, create a data set using the `tibble()` function and save the data set as
a variable called `continent_names`. The data will contains one row for each
continent and should have two variables: `continent` and `continent_fr`. This
gives a mapping between the English and French names.

```{r, question-07}
continent_names <- tibble(
  continent = c("Asia", "South America", "North America", "Africa", "Europe"),
  continent_fr = c("Asie", "Amérique du Sud", "Amérique du Nord", "Afrique", "Europe")
)

continent_names
```

In the code below, join the `cities` dataset with the `continent_names` dataset.
Pipe the data into a plot with longitude on the x-axis and latitude on the
y-axis, coloring the points with the variable `continent_fr`.

```{r, question-08}
cities %>%
  left_join(continent_names, by = "continent") %>%
  ggplot(aes(lon, lat)) +
    geom_point(aes(color = continent_fr))
```

The plot should now show the legend in French. In the code below, recreate the
plot but add better labels for the x-axis, y-axis, and legend. Add a short 
title.

```{r, question-09}
cities %>%
  left_join(continent_names, by = "continent") %>%
  ggplot(aes(lon, lat)) +
    geom_point(aes(color = continent_fr)) +
    labs(x = "Longitude", y = "Latitude", color = "Continent",
         title = "World Cities by Continent")
```

## Semi_join: an example

We have learned about the verb `slice`. There is another variation called 
`slice_sample` that takes a random subset of the data. It takes a named argument
called `n` that tells us how many rows to randomly select from the data. In
the block below, randomly select five rows from the cities dataset (run the
code a few times to verify that the results change):

```{r, question-10}
cities %>%
  slice_sample(n = 5)
```

The `slice_sample` function respects grouping by sampling randomly within each
group. Test this below we randomly picking two cities from each continent:

```{r, question-11}
cities %>%
  group_by(continent) %>%
  slice_sample(n = 2)
```

We are going to do something slightly different that requires the use of a 
filtering join. We want to pick 5 randomly selected countries from the data set
and then return all of the cities that are within those countries. The sampling
should not be weighted by the number of cities within each countries that are
in our data set.

To start, create a table called `selected_countries` that has a single column
called `country` and five rows that have been selected from the countries in 
the cities data. Make sure that you are not over-weighting based on the number
of cities (so China should be just as likely as Angola to be selected).

```{r, question-12}
selected_countries <- cities %>%
  group_by(country) %>%
  summarize() %>%
  slice_sample(n = 5)

selected_countries
```

Now, use a semi-join to select all of the cities that are in these countries
from the cities data set. Pipe the selected cities into a scatter plot with
longitude on the x-axis, latitude on the y-axis, and the points colored by 
country.

```{r, question-13}
cities %>%
  semi_join(selected_countries, by = "country") %>%
  ggplot(aes(lon, lat)) +
    geom_point(aes(color = country))
```

While a bit artificial, this is a good example of how and why you might use
filtering joins. 
