---
title: "Notebook 08 -- Solutions"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)
library(stringi)

theme_set(theme_minimal())
options(pillar.min_character_chars = 15)
options(dplyr.summarise.inform = FALSE)
options(readr.show_col_types = FALSE)
options(ggrepel.max.overlaps = Inf)
options(width = 85L)

food <- read_csv(file.path("data", "food.csv"))
food_prices <- read_csv(file.path("data", "food_prices.csv"))
```

I have set the options `include=FALSE` and `message=FALSE` to avoid cluttering
the solutions with all the output from this code.

# Practice

## Load Datasets

We will work with the largest cities datasets:

```{r, message=FALSE}
cities <- read_csv(file.path("data", "largest_cities.csv"))
```

We will also work with the entire U.S. cities dataset:

```{r, message=FALSE}
us <- read_csv(file.path("data", "us_city_population.csv"))
```

Please refer to notebook02a for more information about these datasets and how
these datasets are organized.

## Creating new numeric features

Using the `cities` dataset, create a new feature named `city_density` by
dividing the city population (`city_pop`) by the city area (`city_area`).
Use the `select` function to select the city name (`name`), population,
area, and density.

```{r question-01}
cities %>%
  mutate(city_density = city_pop / city_area) %>%
  select(name, city_pop, city_area, city_density)
```

The numbers here are very small. What units is the feature `city_density`
currently in? **Answer**: It is given in millions of people per square
kilometer.

Below, modify your code by multiplying the city density by 1000.

```{r question-02}
cities %>%
  mutate(city_density = city_pop / city_area * 1000) %>%
  select(name, city_pop, city_area, city_density)
```

Now what are the units of the city density feature? **Answer**: Thousands
of people per square kilometer.

Now, summarize the average city density by continent.

```{r question-03}
cities %>%
  mutate(city_density = city_pop / city_area * 1000) %>%
  group_by(continent) %>%
  summarize(sm_mean(city_density))
```

Now, below, using the mutate verb, create a plot with city density on the x-axis
and metro density on the y-axis. Use a log scale for the axes and include points
and text repel labels with the city names.

```{r question-04, warning=FALSE}
cities %>%
  mutate(city_density = city_pop / city_area * 1000) %>%
  mutate(metro_density = metro_pop / metro_area * 1000) %>%
  ggplot(aes(city_density, metro_density)) +
    geom_point() +
    geom_text_repel(aes(label = name)) +
    scale_x_log10() +
    scale_y_log10()
```

What two outliers stick out in this plot? **Answer**: Guadalajara and Jinan
have a much higher metro density than would be expect based on the city
density.

## Creating new categorical features

Create a bar plot showing the counts of the the `koppen_code` feature in the
`cities` dataset. Order the groups categories ordered from the biggest to the
smallest. The labels are short so there should be no need to flip the axes in
the plot.

```{r question-05}
cities %>%
  group_by(koppen_code) %>%
  summarize(sm_count()) %>%
  arrange(desc(count)) %>%
  mutate(koppen_code = fct_inorder(koppen_code)) %>%
  ggplot(aes(koppen_code, count)) +
    geom_col()
```

In the code below, create a new feature called `hemisphere` which is equal to
"North" for cities with a latitude above zero and "South" for cities below zero.
Pipe the data to a plot and color the cities by their hemisphere to verify that
your code is correct.

```{r question-06}
cities %>%
  mutate(hemisphere = if_else(lat > 0, "North", "South")) %>%
  ggplot(aes(lon, lat)) +
    geom_point(aes(color = hemisphere))
```

Repeat this process below, but this time create a feature called "quadrant"
that is equal to "NE", "NW", "SE", and "SE" depending on the sign of the
latitude and longitude features. The easiest way to do this is to start by
defining a default value and them fill in the quadrants with four uses of the
`if_else` function. Note: use the `&` operator to combine two conditions
(i.e., `(lat > 0) & (lon > 0)`).

As above, pipe the data into a plot colored by quadrant to verify your work.

```{r question-07}
cities %>%
  mutate(quadrant = "default") %>%
  mutate(quadrant = if_else((lat > 0) & (lon > 0), "NE", quadrant)) %>%
  mutate(quadrant = if_else((lat > 0) & (lon < 0), "NW", quadrant)) %>%
  mutate(quadrant = if_else((lat < 0) & (lon > 0), "SE", quadrant)) %>%
  mutate(quadrant = if_else((lat < 0) & (lon < 0), "SW", quadrant)) %>%
  ggplot(aes(lon, lat)) +
    geom_point(aes(color = quadrant))
```

## Grouped Mutate

To conclude, we will apply a few grouped versions of the mutate verb. To make
sure we understand how this works, take the `us` dataset, group by city, and
compute the maximum value of the population feature for each city.

```{r question-08}
us %>%
  group_by(city) %>%
  mutate(sm_max(population))
```

Note that this added a new column to a copy of the original dataset showing
the largest population for each city at any point in time.

Building off of this approach, construct a scatter plot that shows the
population of cities in 2010 on the x-axis and the percentage of the 2010
population compared to the maximum population of the city from 1790-2010.
Filter to only include cities that are less than 70% of their maximum size in
2010. Use a text repel layer to label each city. (Note: This will require at
least two different calls to the mutate function).

```{r question-09}
us %>%
  group_by(city) %>%
  mutate(sm_max(population)) %>%
  filter(year == 2010) %>%
  mutate(percent_pop_max = population / population_max) %>%
  filter(percent_pop_max < 0.7) %>%
  ggplot(aes(population, percent_pop_max)) +
    geom_point() +
    geom_text_repel(aes(label = city))
```

## Labels and themes

In the code block below, create a visualization of the `cities` data with
city population on the x-axis, city area on the y-axis, points colored by
continent, and the point size given by the cities population density. Include
informative labels (with units, where needed) for the axes and legends, a
title, subtitle, and caption. Set the `theme_sm` theme and add logarithmic
scales for the x and y axes.

```{r question-10, warning=FALSE}
theme_set(theme_sm())

cities %>%
  mutate(city_density = city_pop / city_area * 1000) %>%
  ggplot(aes(city_pop, city_area)) +
    geom_point(aes(color = continent, size = city_density)) +
    scale_x_log10() +
    scale_y_log10() +
    labs(
      x = "Population of City (millions)",
      y = "City Area (square kilometers)",
      color = "Continent",
      size = "Density (1k per sq km)",
      title = "The World's Largest Cities",
      subtitle = "Relationship between Location, Size, and Population",
      caption = "Sources: Wikipedia, U.N., and UNICEF"
    )
```
