---
title: "20. Map Tiles and Points in Polygons"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

```{r, include=FALSE, message=FALSE}
source("start.R")

library(ggmaptile)
```

## Chicago Data

### Load the Data

For the last part of the semester we will be looking at a number of open
data sets from the City of Chicago. We will load in several here:

```{r, message = FALSE}
comarea <- read_sf(file.path("data", "chicago_community_areas.geojson"))
ziparea <- read_sf(file.path("data", "zip_codes.geojson"))
socio <- read_csv(file.path("data", "census_socioeconomic.csv"))
medical <- read_csv(file.path("data", "chicago_medical_examiner_cases.csv.gz"))
```

Today, our focus on the spatial components of the dataset. The spatial data
contains information about zip codes and community areas. Specifically, we
will look at a set of records by the Medican Examiner's office showing
deaths they have investigated:

```{r}
medical
```

Most of these variables should be relatively self-explanatory.

## Advanced Spatial Data

### Fixing Left Key Joins

When working on Project 3, we identified an issue with the way that R works
when joining a spatial dataset and a non-spatial dataset. Unless the spatial
data comes on the left (e.g., first), the spatial data attributes are lost.
This is a problem if you later want to use spatial methods on the data. For
example,

```{r}
medical %>%
  left_join(ziparea, by = c("residence_zip" = "zip"))
```

I have now determined an easy solution. Simply use the function `st_as_sf` after
performing the join. This will recreate the spatial metadata with the correct
projection and spatial bounding coordinates:

```{r}
medical %>%
  left_join(ziparea, by = c("residence_zip" = "zip")) %>%
  st_as_sf()
```

This should help avoid the need to create many temporary datasets or to do
very large joins of the data.

### Creating Spatial Data

The medical records have a longitude and latitude attached to them based on the
reported place of death. We can create a spatial object from this using the
function `st_as_sf`. We will make sure that R knows these are longitude and
latitude coordinates (`crs = 4326`) and that the lon/lat coordinates are not
removed as columns in the data (`remove = FALSE`). After doing so, we can plot
the data using `geom_sf`, which benefits from nice axis labels and the ability
to do coordinate projections.

```{r}
medical %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE) %>%
  ggplot() +
    geom_sf(alpha = 0.01)
```

Using a low value for alpha shows where the majority of deaths reported by
the Medical Examiner's office are located.

### Spatial Joins

The medical examiner data includes someone's residence zip code. Because we
have the spatial data about zip codes, it seems possible to compute those
zip codes the actual death occured in. To do this, we need to use a method
called a *spatial join*. There are many of these, some of which are quite
complex, but the "points in polygon" method here is one of the most common.

The `spatial_join` function takes two datasets and joins them based on their
spatial data. The datasets need to be in the same projection; here we convert
`ziparea` to lat/lon coordinates. If we join the medical data, converted in
spatial points data, into the zip code polygons, we will see which zip area
each case occured in:

```{r, message=FALSE}
medical %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  spatial_join(st_transform(ziparea, 4326)) %>%
  select(residence_zip, zip)
```

Often these are the same, but there are many cases where the two zip codes do
not match, as you will see in the Practice questions.

We can also run the join in the other direction, putting the polygon data on
the left and the point data on the right. This results in a data set that still
contains polygon data.

```{r, message=FALSE}
medical_geo <- medical %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE)

ziparea %>%
  st_transform(4326) %>%
  spatial_join(medical_geo) %>%
  select(residence_zip, zip)
```

I do not understand why this is the default, but notice that the variables in the
second dataset might show up first in the table. You will see that there are some
missing values in the dataset. These are zip codes that have no deaths occuring in
them. By default, `spatial_join` uses a left spatial join; we can turn that off
as follows:

```{r, message=FALSE}
ziparea %>%
  st_transform(4326) %>%
  spatial_join(medical_geo, left = FALSE) %>%
  select(residence_zip, zip)
```

There are many other kinds of spatial joins that we can run in R, but we will not
go into details about them at this time. For example, we can combine two point
datasets by matching up each point in one table with the closest point in the
second table.

### Distances

Let's do one more final spatial manipulation of our data. We will make
a spatial dataset with a single row corresponding to Millennium Park in
downtown Chicago:

```{r}
millennium_park <- tibble(lon = -87.6243107, lat = 41.882612) %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326)
millennium_park
```

The reason for creating this data point is to compute the distance from
all of the medical cases to the center of the park. The computation for
this looks similar to the `st_area` function we saw last time. The
results are returned in metres, which we convert to kilometres by
dividing by 1000.

```{r}
medical_geo %>%
  mutate(distance_mp = as.numeric(st_distance(geometry, millennium_park)) / 1000)
```

One application of this technique is to filter the data to only include points
in a particular part of the world. Here, we look at just the deaths that occured
near the Park:

```{r, warning=FALSE}
medical_geo %>%
  mutate(distance_mp = as.numeric(st_distance(geometry, millennium_park)) / 1000) %>%
  filter(distance_mp < 3) %>%
  ggplot() +
    geom_sf(color = "grey90") +
    geom_sf(color = "orange", data = millennium_park) +
    geom_sf_label(
      label = "Millennium Park",
      color = "orange",
      data = millennium_park,
      nudge_y = -0.003
    )
```

You can see that it roughly maps out a circle (there is a lake to the East
of the city).

### Raster Graphics (Maptiles)

To wrap up our spatial methods, let's look at something completely different.
Rather than working with precise regions (vector data), it can often be useful
to instead plot spatial data over a map. This map could be something similar
to the maps used in Apple Maps or Google Maps. To start, let's take only a set
of points that are very close to the park:

```{r}
medical_mp <- medical_geo %>%
  mutate(distance_mp = as.numeric(st_distance(geometry, millennium_park)) / 1000) %>%
  filter(distance_mp < 1.5)
```

In order to show these over a map, we simply use a scatter plot (NOT `geom_sf`)
on top of a special layer called `stat_maptiles`. Note that the map times need
to go first, otherwise they will be plotted on top of the points. We can set the
opacity (alpha) of the times to make the data easier to see.

```{r, message=FALSE}
medical_mp %>%
  ggplot(aes(lon, lat)) +
    stat_maptiles(alpha = 0.3) +
    geom_point(aes(color = gun_related)) +
    theme_void()
```

There are many other options that you can tweak with the maptiles. See the link on
the course website for more details.
