---
title: "Exam I Review"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)
library(gapminder)

theme_set(theme_minimal())
options(pillar.min_character_chars = 15)
options(readr.show_col_types = FALSE)
options(width = 85L)
options(ggrepel.max.overlaps = Inf)

economics <- mutate(ggplot2::economics, pop = pop / 1000, unemploy = unemploy / 1000)
```

We are going to do two review questions for the exam. If you can do these, you
should be in good shape! For the review we will work with a build-in dataset
called **economics**. It has six features:

**date**       month of data collection
**pce**        personal consumption expenditures, in billions of dollars
**pop**        total population, in millions
**psavert**    personal savings rate (percentage)
**uempmed**    median duration of unemployment, in weeks
**unemploy**   number of unemployed in millions

Let's look at the first few rows:

```{r}
economics
```

Notice that the first feature has a new data type (date). It has some special
properties that we will look at in future classes. For now we can treat it like
any other number.

## Review Question 1

Draw a line plot with date on the x-axis and the personal savings rate on the
y-axis. Color the line in the color "grey75". Add points in the color "#ff9393"
showing the 24 months with the highest average savings rate:

```{r}
lowest_rate <- economics %>%
  arrange(desc(psavert)) %>%
  slice(1:24)

economics %>%
  ggplot(aes(date, psavert)) +
    geom_line(color = "grey75") +
    geom_point(data = lowest_rate, color = "#ff9393")
```


## Review Question 2

Draw a line plot with date on the x-axis that layers two lines: one giving the
population on the y-axis and the other giving the number of unemployed works
on the y-axis. Color the population line in the color "#ff9393" and the
unemployed line in the color "#999932". Set the size aesthetic to both lines 
equal to 1.5 and make the y-axis scale logarithmic (base 10):

```{r}
economics %>%
  ggplot(aes(date, pop)) +
    geom_line(color = "#ff9393", size = 1.5) +
    geom_line(aes(y = unemploy), color = "#999932", size = 1.5) +
    scale_y_log10()
```



