---
title: "Exam III: JSON, Strings, XML"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggrepel)
library(stringi)
library(jsonlite)
library(xml2)
library(httr)

source("cache.R")

theme_set(theme_minimal())
options(pillar.min_character_chars = 15)
options(readr.show_col_types = FALSE)
options(width = 77L)
options(dplyr.summarise.inform = FALSE)
options(ggrepel.max.overlaps = Inf)
Sys.setlocale(locale = "en_US.UTF-8")
httr::set_config(httr::config(ssl_verifypeer = 0L))

theme_set(theme_minimal())
```

## Instructions

Deadline: Monday, 01 November 2021 at 5pm

This exam is the same format as the notebooks we have been completing in class.
In order to complete it correctly, please keep in mind that:

1. In each section there is exactly one object, data set, or visualization that
you need to create. Make sure that tables and graphic prints out in the knit
file.

2. There are nine questions on this exam. Each is worth 10 points. An 
additional 5 points is assigned based on your code formatting across the
entire exam. 

3. You must submit two files: the Rmarkdown file and a knit version of the
file as an HTML document. These must be uploaded to Box by the deadline. Note
that these will not preview correctly on Box, so be careful when uploading.

4. You may use any static resources, such as course notes and external
websites, but you may not discuss the exam with classmates anyone else.

5. I am happy to answer clarifying questions about the exam or to help you
with unexpected R errors. However, I will not answer content-based questions 
after the exam is posted. Note that I may not be able to answer questions 
sent after noon the day the exam is due.

6. The exam should take no more than 3 hours, but you may use as much time 
as you need.

7. Personal computer issues is not an excuse for not completing the exam on
time. As a back up, you can use the computers in the Jepson computer lab or
in the library. 

Good luck!

### Question 1 [Coffee API]

For this question, you need to make a call to a web-based API that returns
data in a JSON format. The data fetches information about popular coffee drinks.
The URL has the following elements:

  protocol:            https
  authority:           api.sampleapis.com
  path:                /coffee/hot
  query parameters:    none (i.e., there are no query parameters)
  
Below, grab the data from this API using the techniques we learned in class,
saving a parsed version of the JSON data as an R object named `obj`. You do
NOT need to print this object out (it is long, so please don't!).

```{r, question-01}

```

### Question 2 [Coffee API]

Create a data tibble in R using the object you produced in the previous question
that contains one row for each coffee drink and three columns: the title of the
drink, the description of the drink, and the id of the drink. Make sure to use
appropriate data types. You do NOT need to store the drink ingredients.

```{r, question-02}

```

### Question 3 [Hourly Forecast]

The code below loads an object that contains saved data from a weather API. It
has an hourly forecast for Richmond, Virginia over one week.

```{r}
hourly <- read_json("data/forecast_rva_hourly.json")
```

In the code below, create a tabular data set from this object that has one row
for each hour of data. It should save the following features: the number of the
reading, the start time, the temperature, the wind speed, and the short
forecast. Make sure to use appropriate data types for each object. Note that
there are many other features, but you do not need to save them. Make sure to
save the data as an R object (you will need it in the next question). 

```{r, question-03}

```

### Question 4 [Hourly Forecast]

Using the data you created above, create a line plot with the forecast id 
number on the x-axis and the temperature reading on the y-axis. 

```{r, question-04}

```

### Question 5 [Hourly Forecast]

Using the data you created above, group the temperature readings by the short
forecast categories and compute the average temperature of each group. Arrange
the categories from the lowest to the highest average temperature. [Keep in mind
that these readings are in Celsius.]

```{r, question-05}

```


### Question 6 [Hourly Forecast]

The wind speeds are given as textual descriptions, which make them hard to work
with as data. In the text below, create a modified version of the hourly 
forecast data that has a new column containing the numeric value of the wind.
A few wind speeds have ranges, from which you can take the first value. Apply
the function `as.numeric` to the result to convert the extracted value to a
proper number.

[Hint: Use the function `stri_extract_first`.]

```{r, question-06}

```

### Question 7 [The Tempest]

The final three questions work with an XML file containing the full text of
Shakespeare's *The Tempest* (I could not resist trying to stick with the weather
theme in these exams). We can read the data in with the following:

```{r}
tempest <- read_xml("data/tempest.xml")
```

In addition to some additional tags containing the acts, scenes, stage
directions and other metadata, the general structure of the data consists of
blocks of text that look like this:

```
<SPEECH>
  <SPEAKER>GONZALO</SPEAKER>
  <TEXT>
    <LINE>Now would I give a thousand furlongs of sea for an</LINE>
    <LINE>acre of barren ground, long heath, brown furze, any</LINE>
    <LINE>thing. The wills above be done! but I would fain</LINE>
    <LINE>die a dry death.</LINE>
  </TEXT>
</SPEECH>
```

In the code below, create a data table that contains one row for each block of
speech (<SPEECH>) and two columns: one for the speaker and one from all of the
text that they spoke. Hint: You do not need to make any use of the <LINE> tag.

Make sure to save the data table as an object as you will need it in the next
question.

```{r, question-07}

```

### Question 8 [The Tempest]

In the code below, start with the data in the previous question and add a column
that counts the total number of words spoken in each speech (<SPEECH>) tag.

```{r, question-08}

```

### Question 9 [The Tempest]

Starting with the data you created in Question 8 (either save the result or copy
your solution again), determine the total number of words spoken by each character
in the text. Arrange the table from the character with the most words to the least
words.

```{r, question-09}

```
