---
title: "Notebook 21 -- Solutions"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---


## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)
library(stringi)
library(lubridate)
library(sf)
library(units)
library(RcppRoll)

theme_set(theme_minimal())
options(dplyr.summarise.inform = FALSE)
options(width = 77L)
options(lubridate.week.start = 1)
Sys.setlocale(locale = "en_US.UTF-8")

sm_centroid <- function(data) {
  suppressWarnings({ z <- st_coordinates(st_centroid(data)) })
  return(tibble(lon = z[,1], lat = z[,2]))
}
```

## French COVID-19

### Load the Data

We will be using the same datasets from the previous notebooks:

```{r, message = FALSE}
dept <- read_sf(file.path("data", "france_departement.geojson"))
pop <- read_csv(file.path("data", "france_departement_population.csv"))
covid <- read_csv(file.path("data", "france_departement_covid.csv"))
```

This time, we will look into the temporal components of the data
and see how they can be integrated into the spatial visualisations.

## Practice

### Plotting the spread of COVID-19 in the Rhône

Plot a line graph showing the number of patients hospitalised (y-axis) over
time (x-axis) for the departement number 69 ("Rhône").

```{r, question-01}

```

Take the plot from the previous question and change the date breaks and labels to label
each month with its full name.

```{r, question-02}

```

Now, we want to compare this to the number of deaths each day. However, the `deceased`
is a total count rather than the new number of deaths. Use the `lag` function to add
a `deceased_new` variable to the data from département number 69:

```{r, question-03}

```

Now, produce a plot that shows the number of newly deceased people in
département 69 over time:

```{r, question-04, message = FALSE}

```

This plot is noisy because of the (thankfully) much lower counts. Modify your
code from above to produce a plot with a 7 day rolling average:

```{r, question-05, message = FALSE}

```

### Days of the Week: Reporting Bias?

Take the covid dataset, group by the date, take the sum (`sum`) of
the variable deceased across all the départements for each day, and
add a variable `wday` giving the day of the week for each row of the
data:

```{r, question-06}

```

Starting with your code from the previous question, compute the change
in the variable `deceased_sum` for each, and compute the average number
of new deaths by the day of the week:

```{r, question-07}

```

What pattern do you see in the data? Do you have any concerns about the reporting
method used in France?

### More Plotting Practice

For a final plot, pick any department and produce an arrow plot that shows the
change in hospitalization numbers over time. Rather than showing every day, just
show the weekly change from Monday to Monday. Color the lines according the direction
of the change and try to include a good scale for the x-axis.

```{r, question-08, warning = FALSE}

```

You can change (hopefully easily) the département number and quickly see the
changes for another region as well.
