---
title: "Notebook 20 -- Solutions"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggrepel)
library(stringi)
library(sf)

source("cache.R")

theme_set(theme_minimal())
options(pillar.min_character_chars = 15)
options(readr.show_col_types = FALSE)
options(width = 77L)
options(dplyr.summarise.inform = FALSE)
options(ggrepel.max.overlaps = Inf)
Sys.setlocale(locale = "en_US.UTF-8")
sm_centroid <- function(data) {
  suppressWarnings({ z <- st_coordinates(st_centroid(data)) })
  return(tibble(lon = z[,1], lat = z[,2]))
}
spatial_join <- function(...) {
  return(st_as_sf(as_tibble(st_join(...))))
}

theme_set(theme_minimal())
```

### Application: French COVID-19 Data

Today, let's look at an application using data from the COVID-19 pandemic in
France. We'll see some more pandemic data in the coming weeks. Today we will
use several data sets. Load them in below and take a moment to look at them.

```{r}
covid <- read_csv("data/france_departement_covid.csv")
pop <- read_csv("data/france_departement_population.csv")
cities <- read_csv("data/france_cities.csv")
geo_fr <- read_sf("data/france_departement.geojson") %>%
  filter(departement <= "95")
```

To get started, produce a basic spatial plot of the `geo_fr` data:

```{r, fig.height=7, question-01}

```

Now, filter the dataset `covid` to the date "2020-05-01" and left join it into
the `geo_fr` dataset by the key "departement". Just have the data print out to
the notebook.

```{r, question-02}

```

Now, starting with the code that you created above, plot a map of the French
Département according to the the number of people that were hospitalized on
1 May 2020. Use `scale_fill_distiller` to pick a good color scale.

```{r, fig.height=7, question-03}

```

A problem with the above map is that it is not normalized by population. Adapt
the code above in the block below by joining in the `pop` dataset and instead 
plotting the hospitalization rate were 100k people.

```{r, fig.height=7, question-04}

```

We also have a data set of French cities called `cities`. Create a spatial
points version of the cities data and save it as an object called `cities_geo`.

```{r, question-05}

```

Now, put together the hopsitalization rate map with the cities data in a single
plot. Include both points and text labels for the cities. 

```{r, fig.height=7, warning = FALSE, question-06}

```

### Application: World Cities

Next, let's load a data set of world cities. This has a large number of cities
and only includes the name, latitude, and longitude.

```{r}
world <- read_csv("data/world_cities.csv")
world
```

Create a new version of the world cities data called `world_geo` by converting
it into a spatial points data set:

```{r, question-07}

```

Perform a spatial join of `world_geo` with `geo_fr`. Just look at the results of
the table without plotting it.

```{r, question-08}

```

Notice above that many of the cities are not in France and therefore do not
match anything in `geo_fr`. To remove these, add the option `left = FALSE` to
the function `spatial_join`:

```{r, question-09}

```

With the data you created in the previous question, plot the French Départment
polygons with the cities (don't include there names; there are too many) plotted
on top:

```{r, fig.height=7, question-10}

```

### US Roads

As a final task, let's read in the US States data again, renaming one of the
variables to make it easier to work with below:

```{r}
state <- read_sf("data/state.geojson") %>% rename(state = name)
state
```

Next, we read in a dataset of four US highways. When loading the data, we will
convert its projection to match that of the state data.

```{r}
hwy <- read_sf("data/usa_roads_95.geojson") %>%
  st_transform(st_crs(state)) 
hwy
```

Create a spatial plot of `hwy`, coloring the lines according to the name of
the highway.

```{r, question-11}

```

Now, perform a spatial join of `hwy` into `state`. Look at the results (do not
yet plot them).

```{r, question-12}

```

Now, take the data you just created and plot the highways again but this time
color by the state abbreviation.

```{r, question-13}

```

As a final task, we want to create a plot that includes both the states and the
highways. However, we only want to include highways that contain some part of
one of the highways in our data. This is slightly more difficult that it may
appear at first, so let's walk through it together.

Start by creating a data set called `hwy_with_state` that joins `hwy` to the
`state` data.

```{r, question-14}

```

Now, we want to create a data set called `state_in_hwy` from a semi join of
`state` with `hwy_with_state`. R won't let us do this directly though because
we are not allowed to do a semi join with two spatial datasets directly.
Instead, we first need to create a non-spatial version of `hwy_with_state`
with this code:

```{r}
hwy_with_state_tbl <- hwy_with_state %>%
  as_tibble() %>%
  select(-geometry)
```

Now, create `state_in_hwy` as a semi join between `state` and
`hwy_with_state_tbl`:

```{r, question-15}

```

Now, create the desired plot using `hwy_with_state` and `state_in_hwy`. Take
a moment to try to make the plot look nice.

```{r, fig.height = 5, question-16}

```


