---
title: "Notebook 18"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggrepel)
library(stringi)
library(jsonlite)
library(xml2)
library(httr)

source("cache.R")

theme_set(theme_minimal())
options(pillar.min_character_chars = 15)
options(readr.show_col_types = FALSE)
options(width = 77L)
options(dplyr.summarise.inform = FALSE)
options(ggrepel.max.overlaps = Inf)
Sys.setlocale(locale = "en_US.UTF-8")

theme_set(theme_minimal())
```

## Review 1: Calling an API

For this question, you need to make a call to a web-based API that returns
data in a JSON format. The data fetches information about popular Beers. The
URL has the following elements:

  protocol:            https
  authority:           api.sampleapis.com
  path:                /beers/stouts
  query parameters:    none (i.e., there are no query parameters)
  
Below, grab the data from this API using the techniques we learned in class,
saving a parsed version of the JSON data as an R object named `obj`. 

```{r, question-01}

```

## Review 2: Parsing JSON

Create a data tibble in R using the object you produced in the previous question
that contains one row for each beer and four columns: the name of the beer, the
price of the beer, the average rating, and the number of ratings. Make sure to
use appropriate data types (though price will need to be a character value for
now). You will want to save the data with a name for use in the next section.

```{r, question-02}

```

## Review 3: Working with Strings

The price in the data you made in the previous section is not a number but a
string because of the dollar sign. Starting with the data in the previous question,
use a regular expression to remove the dollar sign and the function as.numeric
to create a column called `price_num` that is the price in dollars as a number.

```{r, question-03}

```

Now, produce a scatter plot that shows each beer with review average on the
x-axis and price (as a number) on the y-axis.

```{r, question-04}

```

## Review 4: Parsing XML

Below, we read in an XML file taken from the following source (you should take
a look as it shows the structure of the file well):

   https://teibyexample.org/examples/TBED04v00.htm#shakespeare

```{r}
sonnet <- read_xml("data/sonnet17.xml")
```

This is a short file, and we will just use it to practice some XPath expressions.
Below, grab the text of all of the lines (<l>) from the text and store them in
a data table with a single column called "line":

```{r, question-05}

```

Now create a data table that has one row for each foot of the poem; include a
column called "foot" that contains the text, and columns contained "real" and
"met" containing their respective attributes

```{r, question-06}

```

Now, repeat the previous question but only select those segments that are
in the quatrain:

```{r, question-07}

```

Finally, starting with the data created that has one row for each line,
add a column to the data that counts the number of words in each line:

```{r, question-08}

```
