---
title: "Notebook 16 -- Solutions"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.


```{r, include=FALSE, message=FALSE}
source("start.R")
source("cache.R")
```

## Working with strings

### Food Data

To start, as in the notes, we will look at the food data set. We will subset to
only the columns we need for today

```{r}
food <- read_csv("data/food.csv") %>%
  select(wiki, food_group, description)
food
```

In the code block below, create a dataset with one row for each food group that
gives the average number of italics phrases (the `<i>` tag) in descriptions of 
foods within that food group. Arrange the dataset from the highest numbers of 
terms to the smallest. This will take several different verbs to complete.

```{r, question-01}

```

Now, create a dataset with two columns: wiki and italic_phrase that has one row
for each phrase in italics within all of the descriptions. Make sure to not 
include any missing data (so foods without any italics terms will not be
present) and remove the tags themselves in the output data.

```{r, question-02}

```

Create a dataset named `tokens` that has one row for each word in the
descriptions.  Include only two columns: wiki (the id of the food item) and word
(the word itself). Make the words lowercase and do not include extra punctuation
such as commas, periods, and the HTML tags.

Hint: "<[^>]+>" a very helpful regular expression to understand here

```{r, question-03}

```

### Qu'ran Again

We will now read in a complete version of the data you made in the previous
set of notes containing the entire text of the Qu'ran, with one row for each
ayah (verse).

```{r}
quran <- read_csv("data/quran.csv")
quran
```

Translate the code you had in question-03 below to create a tokens dataset for
all tokens in the `quran` data. Include columns for the surah, ayah, and the
word.

```{r, question-04}

```

Create a data set with one row for each surah which pastes together the 7 most
commonly used words in each surah.

```{r, question-05}

```

What do you notice about these terms? Are the very useful for understanding
the texts? **Answer**: 

Now create a dataset named `to_remove` that contains one row for each word that
occurs in more than 100 ayah (verses). These are very common terms that are not
particularly interesting, as you saw above.

```{r, question-06}

```

Now, repeat question-05, but use an anti-join to remove the most common terms
before looking at the most common 7 terms.  

```{r, question-07}

```

Take a moment to look through the data. Does it do a better job revealing the
content of each text?

Repeat what you did above, but now create a list of removed terms by taking 
any term used in more than 10 surah. Compare the output to the one from above
and think about the different views that each list gives.

```{r, question-08}

```



