---
title: "Notebook 22 -- Solutions"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
source("start.R")
source("cache.R")

sm_centroid <- function(data) {
  suppressWarnings({ z <- st_coordinates(st_centroid(data)) })
  return(tibble(lon = z[,1], lat = z[,2]))
}
spatial_join <- function(...) {
  return(st_as_sf(as_tibble(st_join(...))))
}
```

## French COVID-19

### Load the Data

We will be using the same datasets from the previous notebooks:

```{r, message = FALSE}
dept <- read_sf(file.path("data", "france_departement.geojson"))
pop <- read_csv(file.path("data", "france_departement_population.csv"))
covid <- read_csv(file.path("data", "france_departement_covid.csv"))
```

This time, we will look into the temporal components of the data
and see how they can be integrated into the spatial visualisations.

### Plotting the spread of COVID-19 in the Rhône

Plot a line graph showing the number of patients hospitalized (y-axis) over
time (x-axis) for the departement number 69 ("Rhône").

```{r, question-01}

```

Take the plot from the previous question and change the date breaks and labels to label
each month with its full name.

```{r, question-02}

```

Finally, re-create the previous question, changing the labels to be in French.

```{r, question-03}

```

## Chicago Data

### Load the Data

Let's load the Chicago Crime data.

```{r, message = FALSE}
comarea <- read_sf(file.path("data", "chicago_community_areas.geojson"))
ziparea <- read_sf(file.path("data", "zip_codes.geojson"))
socio <- read_csv(file.path("data", "census_socioeconomic.csv"))
medical <- read_csv(file.path("data", "chicago_medical_examiner_cases.csv.gz"))
```

We will look into the temporal components of the data and see how they can be
integrated into the spatial visualisations.

## Practice

### Time of Day: Opioid Related

Let's create a plot that has hour of the day on the x-axis and two lines: one
showing the total number of opiod related deaths in `medical` across an hour 
for every day in the dataset and the other showing the number of non-opiod
related deaths aggregated across each hour. Do this however you would like,
but try to be concise. 

Note: use date_incident_iso as the time variable!

```{r, question-04}

```

Do you notice anything strange? Compute the percentage of non-opioid related
deaths have an incident recorded exactly at midnight (hour and minute are both
0)?

```{r, question-05}

```

Does there seem to be a data issue here?

### Time to Death

Compute the time between the recorded death and the recorded incident. Order
the data with the longest duration at the top. Do you see anything strange?
Is there another data issue here?

```{r, question-06}

```

I want you to investigate the median duration between the time of death and the
time of the incident within the top-100 most common `primary_cause` labels.
Create a table with the median time for each of the top-100 categories and
arrange the table from the shortest to the longest duration

```{r, question-07}

```

Note: If you have the difference in time in seconds, try using the function
`as_hms` to convert these integers into hours, minutes, and seconds. Do the
results seem reasonable to you? Any strange outliers?

### Heat and Cold Deaths

Create a plot with hour on the x-axis and month on the y-axis. Show a scatter
plot with points showing the total number of cold-related deaths that occurred
(incident time, not time of death) at each combination of hour of the day and
month. Try to make the plot look nice (no need for fancy labels, but maybe make
sure to include real month names and properly formatted dates).

```{r, question-08}

```

Repeat the previous question, but look at heat related deaths.

```{r, question-09}

```

Do the patterns surprise you at all?

### Time of the Week

In this final question, I want you to show the number of deaths on the y-axis
as a function of the hour of the week (hour 0 is Monday at midnight, hour 24 is
Tuesday at midnight, etc). It will take a little bit of work to get this
correct, particularly if you want to axes to be labelled correctly without
resorting to hand labels.

```{r, question-10}

```
