---
title: "Notebook 24 -- Solutions"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
source("start.R")
source("cache.R")
```

## Flights

### Datasets

In this notebook we are going to look at a relatively well-known data set of
domestic flights leaving from the three major NYC airports in the year 2013.
There are a few datasets. First, a dataset of the flights themselves. Note that
all of the times are in local time (including arrival hour) and the departure
and arrival delays are in seconds.

```{r}
flights <- read_csv("data/flights.csv")
flights
```

We also have a data set of weather for each hour of the year at the three
airports.

```{r}
weather <- read_csv("data/weather.csv")
weather
```

We also have a data set of the airports. These include the time zone of the 
airport as well as the location as longitude and latitude. We'll use a semi join
to remove small airports that are not in our collection.

```{r}
airports <- read_csv("data/airports.csv") %>%
  semi_join(flights, by = c("faa" = "dest"))
airports
```

Finally, we will also use the state polygons to create a few spatial plots.

```{r}
state <- read_sf("data/state.geojson") %>%
  filter(!(abb %in% c("AK", "HI")))
```

I have already removed the fligths to Alaska and Hawaii, so I also removed 
them above from the state polygons.

### Some Exploratory Analysis

As a warm-up, group the data by the destination airport and compute the number 
of flights to each destination. Arrange the data in descending count.

```{r, question-01}

```

Now, create a spatial plot that shows the location of each airport overlaid over
the state boundaries. Scale the sizes of the airports relative to the number of
flights. Transform the coordinate system to one with crs equal to 5069. 

```{r, question-02}

```

### Departure Time

Now, create a column `dep_time` using the `make_datetime` function from the 
departure time variables. Use the function `force_tz` and the time zone
"America/New_York" to tell R that these are in local NYC time rather than UTC.
Once you get the answer, store the results in the `flights` data.

```{r, question-03}

```

Now, compute the weekday of each flight, group by the weekday, and summarize the
percentage of flights on each weekday that are at least 1 hour late departing.

```{r, question-04}

```

### Weather

We want to join the flights data to the weather data set, group by the variable
`visib` (the visibility in miles), and compute the percentage of flights that
are at least 15 minutes late. Plot the results as a line plot with visib on
the x-axis and the percentage on the y-axis.

```{r, question-05}

```

Repeat the prior question with the wind speed in place of the visibility.

```{r, question-06}

```

### Arrival Time

We now want to create a column `arr_time` as we did above. The arrival times 
are also all local, but are in different time zones. In the code below, join
`flights` to `airports`, create `arr_time` and then use the function `force_tzs`
to convert all of the time zones into "America/New_York". 

```{r, question-07}

```

You may need the help page for `force_tzs` to figure out the syntax. Once you
are happy with the code, save the data back into the variable `flights`.

### Flight Time

Compute the difference between `arr_time` and `dep_time`. Filter the rows where
this difference is negative. There should be a few thousand rows. Look at the
data and try to figure out the problem(s).

```{r, question-08}

```

Update the `flights` data by adding `24 * 60 * 60` seconds to any row where the
arrival time is less than the departure time. This should fix most of the
errors you found above.

```{r, question-09}

```

Take the flights data and compute the difference between the arrival and
departure time. Group by the destination and compute the average flight time.
Join the data with the airports data and arrange from the longest to the
shortest destinations.

```{r, question-10}

```

Take the data you created above and make a spatial plot similar to the first 
one. However, this time we want to color the points according to the average
flight time from NYC. Use the following color scale:

  scale_color_distiller(palette = "Spectral", guide = "legend", n.breaks = 10)
  
Take a moment to adjust the plot to make it look nice. Perhaps set pch = 21?

```{r, question-11}

```


