---
title: "Notebook 12"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)
library(stringi)
library(broom)

theme_set(theme_minimal())
options(pillar.min_character_chars = 15)
options(dplyr.summarise.inform = FALSE)
options(readr.show_col_types = FALSE)
options(ggrepel.max.overlaps = Inf)
options(width = 85L)
```

## Movies Data

Over the next few classes we will be working with a dataset of movies I have
constructed consisting of the top 100 grossing films for each year from 1970 to
2019. The data comes from IMDb. Today we will focus on getting familiar with
the various components of the data. Let's read in the four tables of data,
as well as a data dictionary, and then go through each of the tables.

```{r, message=FALSE}
movies <- read_csv(file.path("data", "movies_50_years.csv"))
m_genre <- read_csv(file.path("data", "movies_50_years_genre.csv"))
m_people <- read_csv(file.path("data", "movies_50_years_people.csv"))
m_dict <- read_csv(file.path("data", "movies_50_years_data_dictionary.csv"))
m_color <- read_csv(file.path("data", "movies_50_years_color.csv"))
```

See the previous notes and the data dictionary for more information about the
available variables.

### Actors and Directors

Summarize the average number of people listed as starring in a film for each
year (first count the number per film and then take the average), and plot the
pattern over the 50 years of data that we have available to us. Do you notice
anything strange about the plot?

```{r, question-01}

```

Filter the data to those names where the gender confidence score is
less than 0.6. Note any patterns that you see and consider caveats that
any gender-based analysis on the larger data set should consider.

```{r, question-02}

```

Now, make a plot showing the number of films starring the 20 most prolific
actors with the bars filled with a color according to an actor's gender. Note,
consider grouping the data by both gender and person before doing the
summarization. 

```{r, question-03}

```

You will (hopefully) notice something strange in the plot above. Fix this by
only including actors with a high gender confidence score (above 0.95 perhaps?).

```{r, question-04}

```


### Three Analyses

We will finish our work with the movies data by replicating three analyses I
did when I was working with this data for my research. 

First, we want to create a table that has one row for each decade in the data.
There is no decade column; you'll need to create it with the code
`floor(year / 10) * 10)`. The columns should provide four summary features: (1)
average brightness of movies from the given decade, (2) average proportion of
the movie posters from this decade that are assigned the color black, (3) average
proportion of the movie posters in this decade code that are assigned the color
white and (4) average proportion of movie posters from this decade that use any
one of the color hues. Include a count variable and sort by the decade.

```{r, question-05}

```

Repeat the previous question but instead of decade, group the data by genre. 
Movies have more than one genre; double count a poster for each genre it is a
part of. Filter the data to include only posters made on or after 1980, 
include only genres with more than 100 films, and arrange by the average amount
of black in the genre.

```{r, question-06}

```

We can associate each film's poster with its most dominant color hue (e.g., not
black, white, grey, or other). Build a table that has one row for each genre
and a column for each color hue (other than 'other') that shows the proportion
of films within each genre that has that color as its dominant hue. Filter the
data to include only posters made on or after 1980, include only genres with
more than 200 films, and arrange by the proportion of films in the genre that 
use the color red.

```{r, question-07}

```



