---
title: "Notebook 22 -- Solutions"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)
library(stringi)
library(lubridate)
library(sf)
library(units)
library(RcppRoll)
library(hms)

theme_set(theme_minimal())
options(dplyr.summarise.inform = FALSE)
options(width = 77L)
options(lubridate.week.start = 1)
Sys.setlocale(locale = "en_US.UTF-8")

sm_centroid <- function(data) {
  suppressWarnings({ z <- st_coordinates(st_centroid(data)) })
  return(tibble(lon = z[,1], lat = z[,2]))
}
spatial_join <- function(...) {
  return(st_as_sf(as_tibble(st_join(...))))
}
```

## Chicago Data

### Load the Data

Let's load the Chicago Crime data.

```{r, message = FALSE}
comarea <- read_sf(file.path("data", "chicago_community_areas.geojson"))
ziparea <- read_sf(file.path("data", "zip_codes.geojson"))
socio <- read_csv(file.path("data", "census_socioeconomic.csv"))
medical <- read_csv(file.path("data", "chicago_medical_examiner_cases.csv.gz"))
```

This time, we will look into the temporal components of the data
and see how they can be integrated into the spatial visualisations.

## Practice

### Time of Day: Opioid Related

Repeat the code from the end of the notes, but now show the number of
incidents (rather than deaths) that occur at each hour of the day, by
whether the incident was opioid related or not.

```{r, question-01}

```

Do you notice anything strange? What percentage of non-opioid related deaths have
an incident recorded exactly at midnight?

```{r, question-02}

```

Does there seem to be a data issue here?

### Time to Death

Compute the time between the recorded death and the recorded incident. Order
the data with the longest duration at the top. Do you see anything strange?
Is there a data issue here?

```{r, question-03}

```

I want you to investiage the median duration between the time of death and the
time of the incident within the top-100 most common `primary_cause` labels.
Create a table with the median time for each of the top-100 categories and
arrange the table from the shortest to the longest duration

```{r, question-04}

```

Note: If you have the difference in time in seconds, try using the function `as_hms`
to convert these integers into hours, minutes, and seconds. Do the results seem
reasonable to you? Any strange outliers?

### Heat and Cold Deaths

Create a plot with hour on the x-axis and month on the y-axis. Show a scatter plot
with points showing the total number of cold-related deaths that occured (incident time,
not time of death) at each combination of hour of the day and month. Try to make the
plot look nice (no need for fancy labels, but maybe make sure to include real month
names and properly formatted dates).

```{r, question-05}

```

Repeat the previous question, but look at heat related deaths.

```{r, question-06}

```

Do the patterns surprise you at all?

### Time of the Week

In this final question, I want you to show the number of deaths on the y-axis as a
function of the hour of the week (hour 0 is Monday at midnight, hour 24 is Tuesday
at midnight, etc). It will take a little bit of work to get this correct, particularly
if you want to axes to be labelled correctly without resorting to hand labels.

```{r, question-07}

```
