---
title: "Notebook 23"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
source("start.R")
source("cache.R")
```

## Keylogs

### Getting Started

Now, let's read in the keylog data that you created. If you have a problem with
your own data, you can use mine (which is already filled in). Take a moment to
see look at the pre-processing that I did to the data. You should be able to
understand most of it. Also, take a moment to look 

```{r}
kl <- read_csv(
  "data/keylogs-1636337989526.csv",
  col_types = "dcccllllldd",
  na = character()
  ) %>%
  mutate(key = if_else(key_code == "Space", " ", key)) %>%
  mutate(key = if_else(key_code == "Enter", "\n", key)) %>%
  filter(type %in% c("up", "down")) %>%
  mutate(action = if_else(stri_length(key) == 1L, "input", "special")) %>%
  select(time, type, key, key_code, action)
kl
```

If we look at just the "down" rows that "input" text, the key values should
more or less spell out all of the text. There will be a few rough edges if you
clicked around or had to use the backspace (but not a problem, I did too!).

```{r}
kl %>%
  filter(type == "down") %>%
  filter(action == "input") %>%
  summarize(key = paste(key, collapse = "")) %>%
  magrittr::use_series(key) %>%
  cat()
```

The last two lines make use of two functions we have not seen so far, but the
general code should be understandable.

### Getting Started

Now for you to do a bit of work using the window functions we saw in the notes.
We will start just looking at the records of when keys were pressed down. In
the code below, filter the data to include only the times that keys were pressed
down, create a variable named `tdiff` that gives the time between the successive
keys, and computes the median differences between these differences. You will
have to use the option `na.rm = TRUE` because there will be a missing value at
the end (or beginning) of the data. This value is in milliseconds.

```{r, question-01}

```

Let us show the distribution of the differences between successive down buttons. 
Recreate the column `tdiff`, filter to have only differences of no more than
400ms, and create a plot of the time differences using `geom_histogram`. Take
a moment to look at the distribution.

```{r, question-02}

```

Finally, let's combine the median difference between down actions based on the
value of the `key` variable. Do this below, arranging the results in descending
order base on the median. Consider adding a count column as well.

```{r, question-03}

```

Take note of the keys that have the largest and smallest median values.

### Grouping Keys

We now want to create a data set that has one row for each key that was pressed
that contains information about when the key was pressed and when it was
released. This can be very tricky if we are not careful, so I will walk you
through the process. Here is a step-by-step description of how we will do this:

- group the data by the `key_code` variable
- create a new variable called `time_down` that is equal to the time 
- create a new variable called `time_up` that gives the time of the next key
- filter to only include rows where the NEXT type is "up"
- filter to only include rows where the type is "down"
- remove rows with a missing time_up
- ungroup the data
- select the columns time_down, time_up, key, key_code, and action

Apply the steps above in the block below. When you are satisfied, save the
results as an object named `keys`.

```{r, question-04}

```

Using the data you just created, compute the median duration between the time
a key was pressed and released.

```{r, question-05}

```
Now, compute the median duration between the time a key was pressed and released
grouped on the value of `key_code`. Arrange the results in descending order of
the median.

```{r, question-06}

```

Take note of the fastest and slowest keys. Do they make sense to you? Finally,
let's compute the duration between the time that a key is released and the time
that the next key is pressed. Compute the median of these differences by the
`key` variable. Arrange in ascending order.

```{r, question-07}

```

Do notice anything interesting? Note that these can be negative. Try to
understand why this is the case.

### Words

As a final task, let's group the data together in order to create a structure
around the individual words. To do this we will end up using a window function.
Before we do the whole thing, let's take a baby step. Below, take the `keys`
data and add a variable called `whitespace` that is a logical value equal to
whether the key is either a space or a newline. Then, add a column called 
`word_id` that is equal to `cumsum(whitespace)`. Look at the results and
understand what it is doing.

```{r, question-08}

```

Below, create a data set called `words` that has the variable `word_id` and 
removes the rows containing whitespae and only includes the action "input".

```{r, question-09}

```

Using the words data, group by the `word_id` variable and compute (within a
summarize verb) three summaries: the minimum value of the time_down, the
maximum value of the time_up, and the value of the word as the pasted together
keys. 

```{r, question-10}

```

Starting with what you did above, compute the time it took you to type each
word. Arrange the data from the longest time to the shortest time. 

```{r, question-11}

```

Finally, if you have time, filter the words to just the first 20 ids. Then, 
create a geom_segment plot showing when each word was typed. Try to add some
labels. 

```{r, question-12}

```

Notice that there are some errors because of mis-typed words (we would need more
complex logic to deal with those).