---
title: "Notebook 25 -- Solutions"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
source("start.R")
source("cache.R")
```

## Italy COVID-19 Data

Today we will look at several datasets concerning Italy and the spread of 
COVID-19 in the country. The main data contains one row for each combination of
province and date. It provides the number of newly diagnosed cases for each
date in a given province.

```{r}
covid <- read_csv("data/it_province_covid.csv")
covid
```

We also have a dataset of the major Italian cities. These contain longitude and
latitude as well as the city's population.

```{r}
city <- read_csv("data/italy_cities.csv")
city
```

Finally, we have a spatial dataset that contains the 107 Italian provinces.

```{r}
province <- read_sf("data/it_province.geojson")
province
```

## Questions

### Cases over Time

Grouping by date, compute the total number of new cases across all provinces.
Create a plot with date on the x-axis and cases on the y-axis.

```{r, question-01}

```

Starting with the code you had above, add a `scale_x_date` that has breaks and
minor breaks every 3 months. Label the date axis with "%Y %b".

```{r, question-02}

```

Modifying the code you had previously, add a rolling mean with `roll_meanr` 
over 14 days. Include two lines: one for the smoothed mean and the a line of
the raw data with a "dotted" linetype a color of "grey75".

```{r, warning = FALSE, question-03}

```

Notice that the data are very noisy when unsmoothed. Let's see if this is a
day of the week effect. Take the covid data, add a day of the week variable
(use a label), group by the day of the week, and compute the average number of
cases on each weekday. Take note of any patterns or differences.

```{r, warning = FALSE, question-04}

```

Returning the last plot you made, create a rolling smoothing line over 7 days
and another over 10 days. Plot both lines with different colors and observe
which one is noisier. Usually we expect longer times to be smoother, but it will
be the opposite here because of the day of the week effect.

```{r, warning = FALSE, question-05}

```

### Italian Map

Next, take the `province` dataset and plot the regions. Transform the data set
to the CRS code 7794 and fill the regions in the color white. 

```{r, fig.height=4, question-06}

```

Create a dataset called `city_geo` that is a spatial version of the `city` 
data. Arrange the dataset in ascenduing order of population (it will make the
next plot look nicer).

```{r, question-07}

```

Starting with the plot of provinces that you had above and add a layer of the
cities on top. Have the size and color change with respect to the population
of the city and use a `scale_color_distiller` to make a nice color scale.

```{r, fig.height=4, question-08}

```

Finally, create a data set called `city_geo_lgl` of the largest ten Italian
cities. Below, add a `geom_sf_label` to the plot you created above with labels
of the city names for these ten cities.

```{r, fig.height=4, question-09}

```


### Window Functions

Compute the floor of each date value in the `covid` dataset. Group by this 
weekly date and summarize the total number of cases across all of the provinces.
Use the `lag` function to create a column called `color` that is "#cc241d" (red)
when the number of cases increases over a week and "#458588" (green) when the
number of cases decreases over a week. Create a plot with the data with date
on the x-axis and the number of cases on the y-axis. Use a geom_segment layer
showing that starts at the number of cases in the previous week and ends at 
the number of cases in the current week. Color the segments with the color
variable you created and set the size to 1.

```{r, warning = FALSE, question-10}

```

Below, again compute the number of cases each week. Filter the data to only
include weeks that are local maxima: the number of cases was lower in the week
before and after a given week. 

```{r, warning = FALSE, question-11}

```

### Spatial Join

Using a spatial join, combine the `city_geo` with the `province` data. Compute
the sum of the population of all cities in each province, convert the data to
a tibble (use `as_tibble`, this removes the spatial metadata) and select just
the `province` and `pop` columns. Save the result as a data set named
`prov_pop`.

```{r, question-12}

```

Finally, use a left join to combine `province` and `prov_pop`. Create a spatial
plot of the populations of the cities in each province using the fill aesthetic
and a `scale_fill_distiller` scale.

```{r, question-13}

```






